---
title: "Poisson regression with offset"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(lmtest)
library(ggplot2)
library(dplyr)
```

## Introduction

With models for count events, we often assume the dependent variable $y_i$ is drawn from a Poisson population with parameter $\lambda_i$, where

$$
P(Y=y_i|\textbf{x}_i)=\frac{e^{-\lambda_i} \, \lambda_i^{y_i}}{y_i \, !} \, \qquad y_i = 0,1,2, \dots
$$

It can be shown that the expected number of events per period or per unit of space (for some given amount of **exposure**) is given by

$$
E[y_i|\textbf{x}_i] = var[y_i|\textbf{x}_i] = \lambda_i = e^{\textbf{x}_i \, \bf{ \beta }}
$$

The most common formulation for $\lambda_i$ is the log-linear model:

$$
\log \lambda_i = \textbf{x}_i \, \bf{ \beta } + \varepsilon_i
$$

Sometimes, it is more relevant to model rates instead of counts. 

* Individuals are not followed the same amount of time.  

* For example, six cases over 1 year should not amount to the same as six cases over 10 years.  

* We measure occurences of events in groups of people that are not of equal size  

* For example, a count of 4 incidents out of 40 students is relatively smaller than a count of 4 out of 20.  

* Generally speaking, observations are different in some dimension (different populations, different geographic sizes) and the outcome is proportional to that dimension.  

  
Hence, instead of the counts

$$
\log \lambda_i = \textbf{x}_i \, \bf{ \beta } + \varepsilon_i \, ,
$$

we take exposure time $t_i$ for the $i$th individual (in other cases, $t_i$ may represent group/classroom size, etc.) and we model occurence rates $\lambda_i/t_i$: 

$$
\log \frac{\lambda_i}{t_i} = \textbf{x}_i \, \bf{ \beta } + \varepsilon_i \, .
$$

Now, because $\log \frac{\lambda_i}{t_i} = \log \lambda_i - \log t_i$, we may re-cast the above equation as 

$$
\log \lambda_i = \log t_i + \textbf{x}_i \, \bf{ \beta } + \varepsilon_i \, .
$$

* $\log t_i$ is the offset.  

* Offset is a predictor variable and its coefficient is fixed at 1.

* For Poisson models, the actual number of events defines the variance, so we often use models with offset instead of models with $\log \frac{\lambda_i}{t_i}$ on the LHS.  

* At the same time both approaches are possible using the `glm()` function (if ratios are used, we need to weight residuals):

    + `glm(Claims ~ District + Group + Age + offset(log(Holders)), family=poisson, data=Insurance)`  
    + `glm(Claims/Holders ~ District + Group + Age, family=quasipoisson, data=Insurance, weights=Holders)`  
    + see [example here for details](https://stats.stackexchange.com/questions/66791/where-does-the-offset-go-in-poisson-negative-binomial-regression)

--- 

## Case study: Lung cancer incidence in Denmark 1968 - 1971 (over 4-year time period)

This data set contains counts of incident lung cancer cases and population size in
four neighbouring Danish cities by age group.

| Variable | Description                                                     |
|----------|-----------------------------------------------------------------|
| city     |  factor with levels Fredericia,   Horsens, Kolding, and Vejl    |
| age      |  factor with levels 40-54, 55-59,   60-64, 65-69, 70-74, and 75 |
| pop      |  the number of inhabitants in a given city                      |
| cases    |   the number of lung cancer cases                               |


### Loading data 

```{r}
lung <- read.table("_dataW09/lung.txt",header=TRUE)
```


```{r}
str(lung)

```

Incident rate is $\lambda = \frac{\textit{cases}_i}{\textit{pop}_i}$ for $i$th city. In Fredericia for people aged 40-45 it is `r lung %>% filter(city=="Fredericia"&age=="40-54") %>% mutate(lambda = cases/pop) %>% select(lambda)`. 


```{r, echo=FALSE}
boxplot(I(2500*cases/pop)~age,data=lung,col="bisque",main="Expected number of lung cancer counts per year per 10.000 people")
```

---

### Poisson regression


We start by a counter-example and we ignore exposure (population size in given age groups)

$$\log(\lambda_i) = \beta_0 + \beta_1 I(\textit{Age55-59}_i) + \beta_2 I(\textit{Age60-64}_i) + \beta_3 I(\textit{Age65-69}_i) + \beta_4 I(\textit{Age70-74}_i)  + \beta_5 I(\textit{Age>74}_i) $$

where $I(.)$ is a indicator function: (1 if TRUE, 0 otherwise) for each range of age, with
Age40-45 is used as baseline.



```{r}
lungmod1 <- glm(cases ~ age, family=poisson, data=lung)
summary(lungmod1)
```


**Interpretation**:


* `exp(2.11021) = 8.24` is the expected count of cancer cases among individuals aged 40 - 54
* `exp(2.1021 - 0.03077) = 8.00` is the expected count of cancer cases among individuals aged 55 - 59
* `exp(-0.0377) = 0.97` is the ratio of the expected counts comparing the 55 - 59 aged group to the baseline group of age `40 - 54`. Also,  `exp(`$\hat{\beta}_1$`)` is the relative rate ratio (IRR).


If we calculate confidence intervals for all ages we find that all contain 0. So, is there any association between cancer and age?

```{r}
confint(lungmod1)
```



### Likelihood Ratio Test

$H_0: \beta_1=\beta_2=\dots=\beta_i$

$H_1: non H_0$

```{r}
lrtest(lungmod1)
```

and hence, we do not reject the hyphotesis of all $\beta_i = 0$, for $i = 1,2,\dots,5$.

---

## How to improve the analysis?


- We have considered the counts of lung cancer cases.
- Each city and age group have different population sizes.


```{r, echo=T}
lung %>% 
  group_by(city) %>% 
  summarize(city_populatio = sum(pop))
#
```


- So far, we have modeled expected counts for each population group, without accounting for exposure.  

- It may be of interest to know the rate per person, per 4 period of observation. On a log-scale, our model is: 

$$\log(\frac{\lambda_i}{\textit{pop}_i}) = \beta_0 + \beta_1 I(\textit{Age55-59}_i) + \beta_2 I(\textit{Age60-64}_i) + \beta_3 I(\textit{Age65-69}_i) + \beta_4 I(\textit{Age70-74}_i)  + \beta_5 I(\textit{Age>74}_i) + \varepsilon_i$$


- All counts are restricted to the same period of 1968 - 1971, the value  $\lambda_i$ are rates `per
4-years'  

- To obtain an easier interpretation of the rates, we can divide by 4 to get rate per person & year. Also, we multiply by 10.000 to get a rate per 10.000 person per year, (i.e. divide by 2500): 

$$\log \lambda_i = \beta_0 + \beta_1 I(\textit{Age55-59}_i) + \beta_2 I(\textit{Age60-64}_i) + \beta_3 I(\textit{Age65-69}_i) + \beta_4 I(\textit{Age70-74}_i)  + \beta_5 I(\textit{Age>74}_i) + \log({\textit{pop}_i/2500}) + \varepsilon_i$$

This can be easily done by

```{r}
lungmod2 <- glm(cases ~ age + offset(log(pop/2500)), 
                family=poisson, data=lung)
summary(lungmod2)
```


- `log(pop/2500)` is the `offset` (try `?offset`)
- The **offset** accounts for the population size, which could vary by age, region, etc ...
- It gives a convenient way to model rates per person-years, instead of modeling the raw
counts.



The inclusion of the offset, implies that the interpretation of the coefficients should be done in terms of $\log(\lambda_i/) - \textit{offset_i}$ (where offset is `log(pop/2500)`).

* Hence in our case, with an offset of `log(pop/2500)`, we should think of the outcome as log rate per 10,000 individuals per year.

**Interpretation**

- $\beta_0$ is the log rate of cancer cases per 10.000 individuals per year in the age group of
40 - 54 (baseline)
- $\beta_1$ is the log relative rate of cancer cases per 10.000 individuals per year comparing
the age group of 50 - 59 to the baseline age group 40 - 5
- $\beta_2$ is the log relative rate of cancer cases per 10.000 individuals per year comparing
the age group of 60 - 64 to the baseline age group 40 - 54


----


### Quick assigment 

Use Poisson regression to model the number of Insurace claims during a given period. Use `Group` /engine capacity/ and car-owner `Age` as regressors. For offset, use the number of insurance-holders per Group.

The data consist of the numbers of policyholders of an insurance company who were exposed to risk, and the numbers of car insurance claims made by those policyholders in the third quarter of 1973.

| Variable | Description                                                                                 |
|----------|---------------------------------------------------------------------------------------------|
| District | factor: district of residence of   policyholder (1 to 4): 4 is major cities.                |
| Group    | an ordered factor: group of car   with levels `<1 litre`, `1 - 1.5 litre`, `1.5 - 2 litre`, `>2 litre`. |
| Age      | an ordered factor: the age of the owner of car insured in 4 groups labelled `<25`, `25 - 29`, `30 -35`, `>35`.    |
| Holders  | numbers of policyholders.                                                                   |
| Claims   | numbers of claims                                                                           |



```{r}
Insurance <- read.csv("_dataW09/Insurance.csv") 
head(Insurance)
str(Insurance)
#
#
# Estimate the Poisson regression model
#
# Produce Confidence intervals and comment on regressor significance.
```
