---
title: "Binary Choice Models with Fixed Effects"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(bife)
```


### Introduction 

In econometrics, fixed effects binary choice models are important tools for panel data analysis. The package `bife` provides a new approach suggested by Stammann, Heiss, and McFadden (2016) to estimate logit and probit panel data models for binary-response variable

$$P(y_{it}=1|\mathbf{x}_{it},\alpha_i) = 
G(\mathbf{x}_{it}\mathbf{\beta}+\alpha_i), \qquad i=1, \dots, N, \quad t=1, \dots , T.$$

* where $G$ is a logistic function or CDF of Normal distribution.  

* The dependent variable $y_{it}$ is binary.   

* Stric exogeneity is assumed: $E(\varepsilon_{it}|\mathbf{X}_i,\alpha_i)=0$.  

* Conditional independence: $y_{i1}, \dots, y_{iT}$ are independent conditional on $(\mathbf{x}_{it},\alpha_i)$.

* For detailed description of the model, see e.g. Wooldridge: Econometric analysis of cross section and panel data (2010), chapters 15.8.2 and 15.8.3.

* Fixed effect is used in econometrician's sense of having a time-constant dummy for each individual. All other parameters in the model are referred to as structural parameters.


* We are primarily interested in estimating the parameters $\mathbf{\beta}$, but the model also includes individual fixed effects $\alpha_i$. We don't make any assumptions about the marginal distribution of $\alpha_i$ or its correlation with the regressors.

The estimator implemented in this package is based on maximum likelihood estimation (ML) of both $\mathbf{\beta}$ and $\alpha_1,\dots,\alpha_N$. 

* `bife()` applies a pseudo-demeaning algorithm proposed by Stammann, Heiss, and McFadden (2016). Its computational costs are lower by orders of magnitude if $N$ is reasonably large.

* Importantly, for FE logit/probit, the ML estimator is not consistent ($N\rightarrow \infty$) and  the problem can be severe if $T$ is small. [Stammann, Heiss, and McFadden (2016)](https://www.econstor.eu/bitstream/10419/145837/1/VfS_2016_pid_6909.pdf). See also see `{bife}` package vignette or Wooldridge (2010) for details. 

* To tackle this problem, `{bife}` provides an analytical and a jackknife bias correction for the structural parameters $\mathbf{\beta}$ and the average partial effects (Hahn and Newey 2004). Package is well suited to analyse big micro-data where $N$ and/or $T$ are large.

Package provides methods to:

  - `bife()`  - estimate binary choice models with fixed effects with/-out bias correction
  - `apeff_bife()`  - compute average partial effects2 with/-out bias correction


(An alternative to full ML estimation of all parameters is a conditional maximum likelihood estimator which conditions out $\alpha_1,\dots,\alpha_N$ and only estimates $\mathbf{\beta}$. It is for example available with `survival::clogit()` and is consistent under the usual regularity conditions. The problem with this estimator is that its computational burden increases dramatically for larger $T$ values and that partial effects cannot be consistently estimated since this would require estimates of $\alpha_1,\dots,\alpha_N$.)





## Example: Hyslop (1999) - Large N

```{r}
dta <- bife::psid
```

The first example is inspired by Hyslop (1999) who analysed the labor force participation of married women in a classical balanced panel. The sample was obtained from the **Panel Study of Income Dynamics** and contains information about $N=$ `r nrow(dta)/length(unique(dta$TIME))` women that were observed over $T=$ `r length(unique(dta$TIME))` years.

| Variable | Description                |
|----------|----------------------------|
| ID       | individual identifier      |
| LFP      | labor force participation  |
| KID1     | # of kids 0-2              |
| KID2     | # of kids 3-5              |
| KID3     | # of kids 6-17             |
| INCH     | income husband             |
| AGE      | age of woman               |
| TIME     | time identifier            |

```{r, warning=FALSE}
ggplot(dta)+geom_boxplot(aes(factor(TIME),AGE, fill=factor(LFP)))
```



To analyse the labor force participation of married women, we specify the following model:

$$LFP_{it}=\beta_1 AGE_{it}+\beta_2(INCH/1000)_{it}+\beta_3 KID1_{it}+\beta_4 KID2_{it}+\beta_5 KID3_{it}+\alpha_i+\varepsilon_{it},$$

where $LFP_{it}$ indicates the labor force participation of a married woman, $AGE_{it}$ refers to the age, $(INCH/1000)_{it}$ is the husbands income in thousand dollars, and the $KID\ast_{it}$ variables refer to the number of kids in a certain age group.


* `bife()`-based estimation:  


```{r}
mod_logit <- bife(LFP ~ AGE + I(INCH / 1000) + KID1 + KID2 + KID3 | ID, data = dta, bias_corr = "ana")
summary(mod_logit)
```

The parameters of binary outcome variables are difficult to interpret quantitatively. In econometrics, partial effects 
$$\frac{\partial Pr(y_{it}=1) }{\partial x_{iij}}$$

are of more interest. Neither `glm()` nor `survival::clogit()` provide a routine to compute partial effects. `{bife}` package provides the function `apeff_bife()` to compute average partial effects based on the estimated model provided by `bife()`. The user simply has to specify which of the variables are discrete and which type of bias correction should be used for the computation of the avarage partial effects. The left column named apeff refers to usual uncorrected average partial effects and the right column named apeff refers to semi-corrected average partial effects following Stammann, Heiss, and McFadden (2016).

```{r}
apeff_bife(mod_logit, discrete = c("KID1", "KID2", "KID3"), bias_corr = "ana")
```

`bife()` also offers the opportunity to estimate fixed effects probit models by specifiying `model = "probit"`.

```{r}
mod_probit <- bife(LFP ~ AGE + I(INCH / 1000) + KID1 + KID2 + KID3 | ID, 
                   data = dta, bias_corr = "ana", model = "probit")
summary(mod_probit)
```

Although the structural parameters are different compared to the logit model due to a different normalization, the average partial effects are similiar:

```{r}
apeff_bife(mod_probit, discrete = c("KID1", "KID2", "KID3"), bias_corr = "ana")
```

---

## Example: ACS PUMS 2014 - Large "T"

The second example is based on a sample drawn from the American Community Survey (ACS PUMS 2014) were the panel structure is slightly different in comparison to the structure used in the section before. Instead of individual fixed effects we consider state fixed effects. $N$ can be now considered as the number of groups (states) and $T_i$ as the group size of group $i$.

| var.  | Description                                                            |
|-------|------------------------------------------------------------------------|
| ST    | state identifier                                                       |
| AGEP  | age of woman                                                           |
| FER   | indicates if a woman gave birth to   a child within the past 12 months |
| PINCP | total persons income                                                   |
| LFP   | labor force participation                                              |


In this example we observe a total of 662,775 married women in $N=51$ states. Since each state is of different population size, we end up with a highly unbalanced panel were the largest state consists of $T_{max}=74,752$ and the smallest of $T_{min}=855$ married women.

The model can be described as follows:

$$LFP_{it}=\beta_1 AGEP_{it}+\beta_2(PINCP/1000)_{it}+\beta_3 FER_{it}+\alpha_i+\varepsilon_{it}$$

where $LFP_{it}$ indicates the labor force participation of a married woman, $AGEP_{it}$ refers to the age, $(PINCP/1000)_{it}$ is the total persons income in thousand dollars, and $FER_{it}$ indicates if a woman gave birth to a child within the past 12 months. In this example $i$ refers to one of the states and $t$ refers to one of the individuals observed in this state.


```{r}
mod_logit <- bife(LFP ~ AGEP + I(PINCP / 1000) + FER | ST, data = acs, bias_corr = "ana")
summary(mod_logit)
```


```{r}
apeff_bife(mod_logit, discrete = "FER")
```

---

This example was based on lecture by [Daniel Czarnowske, Florian Heiss, Amrei Stammann](https://cran.r-project.org/web/packages/bife/vignettes/bife_introduction.html) and slightly modified.