---
title: "Spatial Tobit"
output: html_document
---

```{r setup, include=FALSE}
library(spatialprobit)
library(spatialreg)
```

## General description of a spatial Tobit model

* Spatial Tobit model is defined in terms of a latent dependent variable $z$:

$$\mathbf{z}= \rho \mathbf{Wz} + \mathbf{X} \beta + \mathbf{\varepsilon}, \qquad \varepsilon  \sim N(0, \sigma^2_{\varepsilon} \mathbf{I}_n)$$

* DGP may be written as

$$\mathbf{z} = (\mathbf{I}_n - \rho \mathbf{W})^{-1} \mathbf{X} \beta + (\mathbf{I}_n - \rho \mathbf{W})^{-1} \varepsilon$$

* The observed dependent variable $y$ is censored: 

$$
y_i = \max(z_i, 0)
$$

* Hence, $y$ is only observed for $z \geq 0$ and censored to $0$ otherwise. 

* For Bayesian estimation, Normal distribution is assumed for $\beta$ and Beta distribution is often assumed for $\rho$. [See additional description here](https://www.rdocumentation.org/packages/spatialprobit/versions/0.9-11/topics/sartobit)


--- 

## Spatial Tobit - empirical example (simulated dataset) 


```{r}
set.seed(123456)
# Example from LeSage/Pace (2009), section 10.3.1, p. 302-304
# slight amendment to 'a' and 'beta' is made.
a <- -1.3             # controls degree of censored observation
n <- 1000           # number of simulated "observations"
rho <- 0.7          # spatial dependency parameter (population parameter)
beta <- c(0.4, 2)   # beta_0 and beta_1 parameters (population parameters)
sige <- 0.5         # variance of random element (\sigma^2_{\varepsilon})
I_n <- sparseMatrix(i=1:n, j=1:n, x=1) # I_{1000} identity matrix
x <- runif(n, a, 1) # simulated regressor values
X <- cbind(1, x)    # regressor matrix
eps <- rnorm(n, sd=sqrt(sige)) # random element (a vector)
```

* Define spatial pattern (simulated)

```{r}
# random locational coordinates and 6 nearest neighbors
lat <- rnorm(n)
long <- rnorm(n)
W <- kNearestNeighbors(lat, long, k=6)
```

* Generate $z$ (latent) observations using Spatial Tobit `DGP`:

```{r}
# latent variable z
z <- as.double(solve(I_n - rho * W) %*% (X %*% beta + eps)) 
table(z > 0)
```

* Simulate censoring of observations

```{r}
# full information
yfull <- z
# set negative values to zero to reflect sample truncation
y <- z
ind <- which(y <=0)
y[ind] <- 0
```


* Estimate the Spatial Tobit model 

    + For comparison purposes, `fit_sar` is estimated without censoring of the dependent variable  
    + `fit_sartobit` object contains the estimation based on censored data  

```{r, warning=FALSE}
# Fit SAR (with complete information)
fit_sar <- sartobit(yfull ~ x, W,ndraw=1000,burn.in=200, showProgress=FALSE)
summary(fit_sar)
# Fit SAR Tobit (with approx. 50% censored observations)
fit_sartobit <- sartobit(y ~ x,W,ndraw=1000,burn.in=200, showProgress=F)
summary(fit_sartobit)
```

*  MCMC trace plots

```{r}
par(mfrow=c(2,2))
for (i in 1:4) {
 ylim1 <- range(fit_sar$B[,i], fit_sartobit$B[,i])
 plot(fit_sar$B[,i], type="l", ylim=ylim1, main=fit_sartobit$names[i], col="red")
 lines(fit_sartobit$B[,i], col="green")
 legend("topleft", legend=c("SAR", "SAR Tobit"), col=c("red", "green"), 
   lty=1, bty="n")
}
```

---- 

* Marginal effects for spatial Tobit models (SAR)

```{r}
APE <- marginal.effects(fit_sartobit)
# Direct effects: summary
APE$summary_direct
# Indirect effects: summary
APE$summary_indirect
# Total effects: summary
APE$summary_total
```

-----


