---
title: "Case-control sampling and logistic regression"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
library(dplyr)
library(GGally)
```

### Often, in binary dependent variable models, "success" cases are rare.

* Usually, we take all such recorded "success" cases into our sample,
* Control observations are often limited (say, hundreds),
* As a result, sample and population "success" rates are out of proportion,
* Estimated coefficients and predicted probabilies are affected.
* Special case of non-random sample selection. 



### Usage of case control models

* Rare diseases vs. population (many biostatistics applications), 
* e-shop purchase vs. page-visits (we collect and analyse viewer behavior data).


### Possible approach to Logit estimation & predcting with case control models

All "successful" cases plus approx. 5-times that number of controls is usually sufficient for Logit.

* For predictions, $\hat{\beta}_0$ needs adjustment

    + We control for different population and sample ``success'' rates,
    + case control adjustment formula for the (Intercept)
    + $\hat{\beta}_0^\ast = \hat{\beta}_0 + \log\frac{\pi}{1-\pi}-\log\frac{\tilde{\pi}}{1-\tilde{\pi}}$  
    + where  
    + $\pi$ is the population prevalence ("success" ratio)  
    + $\tilde{\pi}$ is the sample prevalence.  
    + $\hat{\beta}_0^\ast$ is the "true" intercept, relevant for our population  


* $\hat{\beta}_j$ : estimated coefficients for different regressors are not affected

---

# Empirical example: Myocardial infarction (MI) in South Africa

The variables are

|   variable      | description                                   |
|-----------------|-----------------------------------------------|
|   sbp           | systolic blood pressure;                      |
|   tobacco       | cumulative tobacco (kg);                      |
|   ldl           | low densiity lipoprotein   cholesterol;       |
|   adiposity     | a measure of fat;                             |
|   famhist       | family history of heart disease   (Present    |
|   obesity       | BMI                                           |
|   alcohol       | current alcohol consumption;                  |
|   age           | age at onset;                                 |
|   chd           | response  coronary heart disease (MI).        |


## Dataset origin:

Sample of adult males in a heart-disease high-risk region of the Western Cape, South Africa, approx. 1980.

Individuals are admitted to a hospital with MI symptoms (chest-pain, etc.). Upon medical examination, patiens are diagnosed as:

* chd = 1, MI is present 
* chd = 0, MI not present


Sample & population MI ratios do not hold (for obvious reasons):

* There are roughly two negative cases per one positive case of MI (`chd=1`).
* The prevalence rate of MI is 5.1 % in the region.


```{r,warning=FALSE, message=FALSE}
rm(list=ls())
heart <- read.csv("data/SouthAfrica.csv")
head(heart)
str(heart)
summary(as.factor(heart$chd)) # MIs in our sample
plot(heart[ , 1:8], cex=0.8, pch=16, col=as.factor(heart$chd))
```

```{r, warning=FALSE, message=FALSE}
ggpairs(data=cbind(heart[,c("age","tobacco","alcohol","obesity","famhist")],chd=as.factor(heart$chd)), 
        aes(colour = chd), 
        upper = list(continuous = "density", combo="box"))
```


* Our goal is to identify relative strengths and directions of risk factors.
* (Problem: "success" rate in the sample is not representative towards population)
* Although it's representative towards individuals admitted in the hospital with suspected MI.

## Logistic regression

```{r}
heartfit <- glm(chd~., data = heart , family = binomial)
summary(heartfit)
```

With case-control samples, we can estimate the regression
parameters $\beta_j$ accurately - if our model is correctly specified.

The constant term $\beta_0$ is biased.
.. but may be corrected relatively easily


```{r}
pi.pop <- 0.051 # "success" ratio in population
pi.smpl <- 160/462 # "success" ratio in sample
```

case control adjustment formula for the (Intercept)

$$\hat{\beta}_0^\ast = -3.966 + \log\frac{0.051}{1-0.051}-\log\frac{0.346}{1-0.346}$$
where

* $0.051$ is the population prevalence ("success" ratio)
* $0.346$ is the sample prevalence.
* $\hat{\beta}_0^\ast$ is the "true" intercept, relevant for our population


```{r}
heart.coeffs <- heartfit$coefficients
heart.coeffs
(beta.0.pop <- heart.coeffs[1] + log(pi.pop/(1-pi.pop)) - log(pi.smpl/(1-pi.smpl)))
```

## Logit coefficients adjusted for case control estimates:

```{r}
heart.coeffs[1] <- beta.0.pop
heart.coeffs
length(heart.coeffs) # Intercept + 8 variables
```


## Predictions made from the estimated model:

Lets calculate MI probability for a *typical* middle-age SA male with family history of MI and for "the same" male without family history of MI 

* Note that our typical male is constructed by averaging sample characteristics.  
* However, the question is: If a male of `age` = x, with `sbp` = y, ... is drawn randomly from a **population**, what is his probability of MI in a given year?  
* Characteristics of a "representative" individual can be set as necessary (shown next). 


```{r}
attach(heart) 
# We append two rows at the end of "heart" dataframe:
# first is for typical male with MI family history "Present"
# second is for typical male with MI family history "Absent"
# note the "NA" values in the "chd" column 
#
heart2 <- rbind(heart, list(mean(sbp), mean(tobacco), mean(ldl), mean(adiposity), "Present", mean(obesity),
                 mean(alcohol), mean(age), NA))
heart2 <- rbind(heart2, list(mean(sbp), mean(tobacco), mean(ldl), mean(adiposity), "Absent", mean(obesity),
                            mean(alcohol), mean(age), NA))
tail(heart2)
```



Predictions based on **sample** "success" prevalence (beta 0 is biased) -- no case-control correction (biased):

```{r}
# Estimated model "heartfit" and newdata are combined here:
predict(heartfit, newdata=heart2[463:464, ], type="response")
```

## Case-control: corrected predictions

```{r}
#  This prediction is done "manually", using the amended
#  heart.coeffs  vector of coefficients
#  and a vector of "observed" predictors (average males)
#  .. However, we cannot multiply Factors....
#
heart3 <- heart2     # We prepare yet another dataframe 
heart3$famhist <- ifelse(heart3$famhist == "Present",1,0)
# First, we calculate log-odds for the two "typical males":
(Log.Odds1 <- heart.coeffs[1] + sum(heart3[463, 1:8]*heart.coeffs[2:9]))
(Log.Odds2 <- heart.coeffs[1] + sum(heart3[464, 1:8]*heart.coeffs[2:9]))
```

Expected probability of MI:

```{r}
MI.prob.Avg.male.Famhist.Yes <- exp(Log.Odds1)/(1+exp(Log.Odds1))
MI.prob.Avg.male.Famhist.no <- exp(Log.Odds2)/(1+exp(Log.Odds2))
names(MI.prob.Avg.male.Famhist.Yes) <- "Expected MI probability: average male WITH family history of MI"
names(MI.prob.Avg.male.Famhist.no) <- "Expected MI probability: average male WITHOUT family history of MI"
MI.prob.Avg.male.Famhist.Yes
MI.prob.Avg.male.Famhist.no
```

---

## Assignment
```{r}
summary(heart[,c("tobacco","age","ldl")])
```
Calculate MI probability for a male with major risk factors in the fourth quartile,

* set tobacco = 8
* set age = 60
* set ldl = 5.790 (anything above 4.12 is considered highly risky)
* calculate for both family history "Present" and "Absent"  
* The simples way to proceed with this assignment is to amend `R` code starting on line `[157`].  
* $\dots$ (You may start by making a copy of this Rmd before amending the code)