---
title: "Spatial probit"
output: html_document
---

```{r setup, include=FALSE}
library(Matrix)
library(sf)
library(osmdata)
library(spatialprobit)
library(spatialreg)
```

## General description of a spatial probit model

* We have the spatial autoregressive model (SAR model, spatial lag model)

$$
z = \rho W z + X \beta + \varepsilon, \qquad \varepsilon \sim N(0, \sigma^2_{\varepsilon} I_n)
$$
* and the corresponding DGP for $\mathbf{z}$ is:

$$
z = (I_n - \rho W)^{-1} X \beta + (I_n - \rho W)^{-1} \varepsilon
$$

In a spatial probit model, $z$ is regarded as a latent variable, which cannot be observed. Instead, the observables are only binary variables $y_i$ (0, 1) as

$$
y_i = 
\begin{cases}
1 & \quad \text{if}~z_i \geq 0, \\
0 & \quad \text{if}~z_i < 0.
\end{cases}
$$

---


* The model is estimated using a Bayesian approach: [Estimating Spatial Probit Models in R](https://journal.r-project.org/archive/2013/RJ-2013-013/RJ-2013-013.pdf)  

    + $\beta$ is assumed Normally distributed (with suitable prior parameter values),  
    + $\rho$ is often assumed to follow Gamma distribution,  
    + $\sigma^2_{\varepsilon} = 1$ is often assumed (for identification).



* Marginal effects (direct, indirect, total) are calculated from an estimated model.

* For some $r$th regressor $x$, we can write   

$$
\mathbf{S}_r(\mathbf{W}) = \frac{\partial E [\mathbf{y}|\mathbf{x}_r]}{\partial \mathbf{x}_r^{\prime}} = \phi\left(\left[(\mathbf{I}_n -  \rho \mathbf{W})^{-1} \bar{\mathbf{x}}_r \beta_r \right] \odot (\mathbf{I}_n -  \rho \mathbf{W})^{-1} \mathbf{I}_n \beta_r \right)
$$
where $\mathbf{S}_r(\mathbf{W})$ is an $(N \times N)$ matrix with dimension given by the number of spatial units, and the $\odot$ symbol stands for Hadamard product (element-wise matrix multiplication). 

* The direct impact of a change in $x_{ir}$ on "own" observation $y_i$ can be written as the $\{i,i\}$ element of $\mathbf{S}_r(\mathbf{W})$:

$$
\frac{\partial y_i}{\partial x_{ir}} = \mathbf{S}_r(\mathbf{W})_{ii}
$$

* Similarly, the indirect impact from observation $j$ to observation $i$ is

$$
\frac{\partial y_i}{\partial x_{jr}} = \mathbf{S}_r(\mathbf{W})_{ij}
$$

* For detailed description of marginal effects, see [text here](https://www.rdocumentation.org/packages/spatialprobit/versions/0.9-11/topics/marginal.effects)


--- 


## Hurricane Katrina (August 2005) example

* New Orleans business recovery in the aftermath of Hurricane Katrina  

* Probability of business recovery (store reopening after the flood) modeled as a function of flood depth, store size, store ownership, etc.  

* Different time horizons for the reopening can be considered (0-3 months, 0-6 months, etc.)


#### Data 

The dataset contains 673 observations on 3 streets in New Orleans and can be used to estimate the spatial probit models and to replicate the findings in the paper.

| Variable  | Description                                                            |
|-----------|------------------------------------------------------------------------|
| long  | longitude coordinate of store     |
| lat   | latitude coordinate of store      |
| flood_depth | flood depth (measured in feet)  |
| log_medinc  | log median income             | 
| small_size | binary variable for "small size" firms   |
| large_size | binary variable for "large size" firms   |
| low_status_customers   |  binary variable for low socio-economic status of clientele  |
| high_status_customers  | binary variable for high socio-economic status of clientele  |
| owntype_sole_proprietor  | a binary variable indicating "sole proprietor" ownership type  |
| owntype_national_chain   | a binary variable indicating "national_chain" ownership type   |
| y1  | reopening status in the very short period 0-3 months; 1=reopened, 0=not reopened  |
| y2  | reopening status in the period 0-6 months; 1=reopened, 0=not reopened  |
| y3  | reopening status in the period 0-12 months; 1=reopened, 0=not reopened |  


```{r, message=FALSE}
data(Katrina)
str(Katrina)
Katrina_sf <- st_as_sf(Katrina, coords = c("long", "lat"), crs = 4326) 

```

#### Map of the area analyzed

```{r, message=FALSE}
mapview::mapview(Katrina_sf)
```

* Using the data, we replicate some output from the paper: J. P. LeSage, R. K. Pace, N. Lam, R. Campanella and X. Liu (2011), New Orleans business recovery in the aftermath of Hurricane Katrina Journal of the Royal Statistical Society A, 174.

* [https://www.rsgis.envs.lsu.edu/docs/LeSage-Journal%20of%20Royal%20Statistics-2011.pdf](https://www.rsgis.envs.lsu.edu/docs/LeSage-Journal%20of%20Royal%20Statistics-2011.pdf)

```{r}
# 0-6 months time horizon
# LeSage et al. (2011) use k=11 nearest neighbors in this case
nb <- knn2nb(knearneigh(cbind(Katrina$lat, Katrina$long), k=11))
listw <- nb2listw(nb, style="W")
W1 <- as(as_dgRMatrix_listw(listw), "CsparseMatrix")
nb
```

Model is estimated using the Bayesian approach

```{r, warning=F}
fit1 <- sarprobit(y2 ~ flood_depth + log_medinc + small_size + large_size +
  low_status_customers +  high_status_customers + 
  owntype_sole_proprietor + owntype_national_chain, 
  W=W1, data=Katrina, ndraw=600, burn.in = 100, showProgress=F)
summary(fit1)
```

Estimate marginal effects (average direct, indirect and total impacts) for the SAR probit

```{r, warning=F}
# replicate LeSage et al. (2011), Table 4, p.1018
# SAR probit model effects estimates for the 0-3-month time horizon
impacts(fit1)
```

-----

#### Quick exercise

* use $k = 15$ for the $kNN$ method.  

* Produce marginal/partial impacts estimates