---
title: "Panel data models and tests"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plm)
library(lmtest)
library(sandwich)
library(car)
library(ggplot2)
```

## Illustrative dataset

* Cornwell and Rupert, Labor Market Data, 595 Individuals, 7 years, balanced panel  
* Analysis focuses on returns to schooling and work experience
* Heads of households form the sample
* For additional description, see Greene, Econometric Analysis, ch. 11.

| Variable | Description                                                          |
|----------|----------------------------------------------------------------------|
| EXP |  Work experience  |
| WKS |  Weeks worked  |
| OCC |  Occupation: 1 if blue collar  |
| IND |  1 if manufacturing industry  |
| SOUTH |  1 if resides in south  |
| SMSA |  1 if resides in a city (SMSA)  |
| MS |  1 if married  |
| FEM |  1 if female  |
| UNION |  1 if wage set by union contract  |
| ED |  Years of education - time invariant (no students in sample) |
| BLK |  1 if individual is black  |
| LWAGE|  Log of wage |

```{r}
WageDF <- read.csv("_dataW07/TableF81.csv")
# Create cross-sectional and time IDs
WageDF$CSID <- rep(1:595,each=7)
WageDF$CSID <- as.factor(WageDF$CSID)
WageDF$YEAR <- rep(1976:1982,595)
str(WageDF)
```

--- 

## Testing individual effects by F-test

* Null hypothesis of common intercept is tested against the alternative of individual-specific intercepts.

* Common slopes are assumed (not tested)

* Unrestricted model: $y_{it} = \beta_{0} + {\mathbf d'\delta}_{0} + \beta_1 x_{it1} + \beta_2 x_{it2} + u_{it}$  
where $\mathbf{d}$ is a vector of CSID-based dummy variables and $\mathbf{\delta}_{0}$ is a vector of regression coefficients.

* Restricted model: $~~y_{it} = \beta_{0}~ \, + \beta_1 x_{it1} + \beta_2 x_{it2} + u_{it}$.


* Can be implemented as an $F$-test for linear (zero) restrictions: Pooled regression vs LSDV model

* Illustrative example model:  `LWAGE` $\leftarrow$  `ED`,`EXP`

```{r}
Pooled.mod <- lm(LWAGE~ED+EXP,WageDF)
summary(Pooled.mod)$coefficients
LSDV.mod <- lm(LWAGE~ED+EXP+CSID,WageDF)
# Show only first few a_i estimates
summary(LSDV.mod)$coefficients[1:10,]
```

* Now, we perform the $F$-test

```{r}
anova(Pooled.mod, LSDV.mod)
```

--- 

## Testing for poolability - Chow test for identical slopes


* Chow tes for identical slopes

* Allow for different intercepts & test for equal slopes in all CS-units

  + Estimate model separately for each CS unit (alt: time period). 

  + Compare with "FE" model (individual intercept, common slopes on regressors) using an $F$-test -- are the slopes identical among CS-units?


* Drawback: test does not work with time-invariant regressors -- as model is estimated individually for each CS-unit, such regressors are perfectly correlated with the intercept; also time-invariant regressors are not compatible with FE estimation.

* `pooltest()` from the `{plm}` package  

* The following "poolability test" is illustrative only (we drop `ED` from regression, as it is time-invariant in our dataset).

* Unrestricted model: $y_{it} = \beta_{0} + \beta_{i1} x_{it} + \mu_i + \nu_{it}$

* Restricted model: $~~y_{it} = \beta_{0}~ \, + \beta_1 x_{it} + \mu_i + \nu_{it}$

    + $H_0:~\beta_{11}=\beta_{21}=\dots=\beta_{N1}$  
    + $H_1:~\neg H_0$


```{r}
# 1 Unrestricted model with variable slopes 
# .. and intercepts for each CS unit
pvcm1 <- pvcm(LWAGE~EXP, data=WageDF, index = c("CSID","YEAR"), model = "within")
summary(pvcm1)
# 2 FE model (individual intercepts, common slopes)
FE.1 <- plm(LWAGE~EXP, data=WageDF, index = c("CSID","YEAR"), model = "within")
summary(FE.1)
#
# Test 1: Null hypothesis is that the dataset is poolable 
#         ..(i.e. individuals have the same slope coefficients)
pooltest(FE.1,pvcm1)
```


  * Alternatively, compare with a "pooling" model -- are all coefficients identical among CS-units (i.e. intercept and slopes)
  
* Unrestricted model: $y_{it} = \beta_{0} + \beta_{i1} x_{it} + \mu_i + \nu_{it}$

* Restricted model: $~~y_{it} = \beta_{0}~ \, + \beta_1 x_{it} ~~~~~~~+ \nu_{it}$ 

    + $H_0:~\beta_{11}=\beta_{21}=\dots=\beta_{N1}$ and $\mu_i = 0, \forall i$
    + $H_1:~ \neg H_0$



```{r}
# 3 Pooled model (common intercept & common slopes)
Pool.1 <- plm(LWAGE~EXP, data=WageDF, index = c("CSID","YEAR"), model = "pooling")
summary(Pool.1)
# Test 2: Null hypothesis is that the dataset is poolable 
#         ..(i.e. individuals have the same intercept and slope coefficients)
pooltest(Pool.1,pvcm1)
```


--- 


## Testing for individual and time effects (Honda, 1985)

* Using OLS-based ("pooling") residuals, we test the null hypothesis of redundant individual $(\mu_i)$ and/or time $\lambda_t$ effects.

    + Individual effects: $y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \mu_i + \nu_{it}$  
    + Time effects: $~~~~~\, y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \lambda_t + \nu_{it}$  
    + Twoways effects: $y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \mu_i + \lambda_t + \nu_{it}$  


* `plmtest()` from the `{plm}` package  

* LM-based tests, use only the residuals of the pooling model (if performed on RE of FE model ("within","random"), corresponding pooling model is calculated internally first).  

```{r}
# Start by estimating a "pooling" model
plm.pool <- plm(LWAGE~ED+EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "pooling")
summary(plm.pool)
```

* Honda (1985) LM test

```{r}
# individual
plmtest(plm.pool, effect= "individual", type="honda")
# time
plmtest(plm.pool, effect= "time", type="honda")
# twoways
plmtest(plm.pool, effect= "twoways", type="honda")
```


--- 


## F-test of individual and time effects: FE vs pooling

* By comparing residuals from the FE (`within`) model and pooled (OLS-estimated) model, we test the null hypothesis of redundant individual $\mu_i$ and/or time $\lambda_t$ effects (given the specification of the `within` model).

    + Individual effects: $y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \mu_i + \nu_{it}$  
    + Time effects: $~~~~~\, y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \lambda_t + \nu_{it}$  
    + Twoways effects: $y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} + \mu_i + \lambda_t + \nu_{it}$  
    + Pooling: $~~~~~~~~~~~y_{it} = \beta_{0} + \beta_{1} x_{it1} + \dots + \beta_k x_{itk} +  \nu_{it}$  
    
    
* Test for individual effects    

```{r}
# Pooled specification, interaction ED*EXP used
WageDF$ED_EXP <- WageDF$ED * WageDF$EXP
plm.pooling <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "pooling")
summary(plm.pooling)$coefficients
# FE model, individual effects, interaction ED*EXP used
plm.FE.ind <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "individual")
summary(plm.FE.ind)$coefficients
# F-test for individual effects
pFtest(plm.FE.ind,plm.pooling)
``` 

* Tests for time and twoways effects    

```{r}
# FE model, time effects, interaction ED*EXP used
plm.FE.time <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "time")
# FE model, twoways effects, interaction ED*EXP used
plm.FE.twoways <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "twoways")
#
# F-test for time effects
pFtest(plm.FE.time,plm.pooling)
# F-test for twoways effects
pFtest(plm.FE.twoways,plm.pooling)
``` 


---


## Hausman test FE vs RE estimator

*  Hausman test is based on the comparison of two sets of estimates 

* `phtest()` from the `{plm}` package  

  + Two main arguments to the test: FE-estimated and RE-estimated models

```{r}
# FE-estimated model, individual effects, interaction ED*EXP used
FE.mod <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "individual")
# RE-estimated model, individual effects, interaction ED*EXP used
RE.mod <- plm(LWAGE~EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "random", effect = "individual")
#
summary(FE.mod)$coefficients
summary(RE.mod)$coefficients
# Hausman test
phtest(FE.mod,RE.mod)
```


--- 

## Wooldridge test: FD vs FE

* $y_{it} = \alpha + \beta_1 + x_{it1} + \dots + \beta_k x_{itk} + \mu_i + \nu_{it}$

* Serial correlation test that can be used as a specification test to choose the most efficient estimator -- FD vs FE.

* If $\nu_{it}$ are not serially correlated,  
  + FE is more efficient than FD  
  + Residuals in the FD model: $e_{it} \equiv \nu_{it}-\nu_{i,t-1}$ are correlated
  
* If $\nu_{it}$ follow a unit root:  
  + FD is more efficient than FE  
  + Residuals in the FD model: $e_{it} = \nu_{it}-\nu_{i,t-1}$ are not serially correlated  


```{r}
# H0 of no serial correlation in FD residuals - default "h0"
pwfdtest(LWAGE~EXP+ED_EXP, data=WageDF, index = c("CSID","YEAR"))
# H0 of no serial correlation in FE residuals
pwfdtest(LWAGE~EXP+ED_EXP, data=WageDF, 
         index = c("CSID","YEAR"),h0="fe")
```

* If both rejected, whichever estimator is chosen will have serially correlated errors: use the autocorrelation-robust covariance estimators.

---- 

## Breusch-Godfrey test for panels 

* Mainly for RE (and pooling) models.  

* Applicable to FE models only if $T$ is large

* `pbgtest()` is a wrapper for `bgtest()` from the `{lmtest}` package.  

```{r}
# The following example serves for pbgtest() illustration only 
# .. RE assumptions not really met.
# RE-estimated model
RE2.mod <- plm(LWAGE~WKS+EXP+ED+MS, data=WageDF,
              index = c("CSID","YEAR"), 
              model = "random", effect = "individual")
#
pbgtest(RE2.mod)
# We can test for ar(2) serial correlation in residuals
pbgtest(RE2.mod, order=2)
```

---- 

## Wooldridge test for short FE panel models 

```{r}
# FE-estimated model
FE2.mod <- plm(LWAGE~WKS+EXP+ED_EXP, data=WageDF,
              index = c("CSID","YEAR"), 
              model = "within", effect = "twoways")
#
pwartest(FE2.mod)
```

---- 

## Autocorrelation-robust (and HAC) covariance estimators  
`
* The main use of `vcovHC` is together with testing functions from the `{lmtest}` and `{car}` packages.

* Robust estimators of the covariance matrix of coefficients are provided

* All types assume no correlation between errors of different CS-units (groups) while allowing for heteroskedasticity across CS-units (groups)

* `"white1"` allows for general heteroskedasticity but no serial correlation.

* `"white2"` is `"white1"` restricted to a common variance inside every CS-unit (group).

* `"arellano"` allows a fully general structure w.r.t. heteroskedasticity and serial correlation.

```{r}
# FE model
FE3.mod <- plm(LWAGE~WKS+EXP+ED_EXP, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "individual")
# compare non-robust and robust estimates
coeftest(FE3.mod)
coeftest(FE3.mod, vcov = vcovHC(FE3.mod, method = "arellano"))
```


--- 

## Tests for cross-sectional dependence (XSD)

* Analogous yet distinct to the more familiar issue of serial correlation.

* Can arise, e.g., if individuals respond to common shocks or if spatial diffusion processes are present, relating individuals in a way depending on a measure of distance (spatial models)
    
* If XSD is present, the consequence is, at a minimum, inefficiency of the
usual estimators and invalid inference when using the standard covariance matrix.

* In `{plm}`, only misspeciffication tests to detect XSD are available -- no robust method to perform valid inference in its presence.

```{r}
# FE model
FE4.mod <- plm(LWAGE~WKS+EXP+ED_EXP+MS, data=WageDF, 
                index = c("CSID","YEAR"), model = "within", effect = "twoways")
pcdtest(FE4.mod)
```

---

