---
title: "Hausman and Taylor model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plm)
data("Wages", package = "plm")
Wages.df <- pdata.frame(Wages, index=595)
```

## Hausman and Taylor (HT) model

* Panel data model, expansion and modification to the RE/CRE approach.

* Time varying and time-invariant regressors are present

* Some regressors are correlated with unobserved individual effects



#### General HT model specification:

$$ 
y_{it} = \mathbf{x}^{\prime}_{1it}\beta_1 + \mathbf{x}^{\prime}_{2it}\beta_2 
       + \mathbf{z}^{\prime}_{1it}\alpha_1 + \mathbf{z}^{\prime}_{2it}\alpha_2
       + \mu_i + \varepsilon_{it}
$$
  
* where
  
    + $\mathbf{x}_{1}$ contains $K_1$ variables: *Time-varying*, **uncorrelated** with $\mu_i$.  
    + $\mathbf{x}_{2}$ contains $K_2$ variables: *Time-varying*, **correlated** with $\mu_i$.
    + $\mathbf{z}_{1}$ contains $L_1$ variables: *Time-invariant*, **uncorrelated** with $\mu_i$. 
    + $\mathbf{z}_{2}$ contains $L_2$ variables: *Time-invariant*, **correlated** with $\mu_i$.   
    + $\beta_1$,  $\beta_2$, $\alpha_1$ and $\alpha_2$ are vectors of coefficients (to be estimated)    

----  

  
* HT model (and estimation method) assumptions:

    + $E[\mu_i|\mathbf{x}_{1},\mathbf{z}_{1}] = 0$;   
    + $E[\mu_i|\mathbf{x}_{2},\mathbf{z}_{2}] \neq 0$;   
    + $\text{var}[\mu_i|\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}] = \sigma^2_{\mu}$;
    + $\text{cov}[\mu_i,\varepsilon_{it}|\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}] = 0$;
    + $\text{var}[\mu_i+\varepsilon_{it}|\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}] = \sigma^2 = \sigma^2_{\mu} + \sigma^2_{\varepsilon}$;
    + $\text{cor}[\mu_i+\varepsilon_{it}, \mu_i+\varepsilon_{is} |\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}] = \rho = \sigma^2_{\mu} / \sigma^2$.  
$~~$  

* OLS does not converge (asymptotically biased, inconsistent) because of endogenous regressors.  

* $\beta_1$ and $\beta_2$ parameters may be consistently estimated by FE/LSDV 

* We do not need instruments for $\mathbf{z}_{1}$ as it is not correlated with $\mu_i$

* $\overline{x}_{1i}$ is a valid instrument for $\mathbf{z}_{2}$ (Hausman and Taylor, 1981)  


--- 

## HT Estimation: step-by-step

1. Estimate $\beta_1$ and $\beta_2$ parameters using the FE (within) estimator


    + $(y_{it}-\overline{y}_i) = (\mathbf{x}_{1it}-\overline{\mathbf{x}}_{1it})\beta_1 + (\mathbf{x}_{2it}-\overline{\mathbf{x}}_{2it})\beta_2 + (\varepsilon_{it}-\overline{\varepsilon}_i)\\$.  
    
    
2. From step 1, use the residuals to compute the within means of residuals


    + $\overline{e}_i = \frac{1}{T} \sum_{t=1}^T e_{it}\\$ and generate
    + $\overline{\mathbf{e}}^{\prime} = (\overline{\mathbf{e}}_1, \overline{\mathbf{e}}_2, \dots, \overline{\mathbf{e}}_n)^{\prime}\\$. 
    
    
3. 2SLS


    + Individually regress all $\mathbf{Z}_{2}$ elements (regressors correlated with $\mu_i$) on $\mathbf{Z}_{1}$ and $\mathbf{X}_{1}$.  
    
    + Using fitted $\hat{\mathbf{Z}}_2$, form matrix $\mathbf{Z}^{*} = [\mathbf{Z}_1,\hat{\mathbf{Z}}_2]$.   
 
    + Regress  $\overline{\mathbf{e}}$ on $\mathbf{Z}^{*}$ to get $\hat{\alpha}_1$ and $\hat{\alpha}_2\\$.


4. Estimate variances $\sigma^2_{\mu}$ and $\sigma^2_{\varepsilon}$ 

    + Use $\hat{\sigma}^2_{\varepsilon}$ from Step 1 (FE)  
    
    + Use the estimated $\hat{\sigma}^2$ from Step 3 (2SLS) to calculate $\hat{\sigma}^2_{\mu}$:
    
    + $\hat{\sigma}^2_{\mu} = \hat{\sigma}^2 - \frac{\hat{\sigma}^2_{\varepsilon}}{T}\\$  


5. Estimate $\hat{\theta}$ for FGLS:

    + $\hat{\theta} = \sqrt{\frac{\sigma^2_{\varepsilon}}{\sigma^2_{\varepsilon}+T\sigma^2_{\mu}}}\\$  
    
    
6. FGLS estimate $\hat{\beta}$ and $\hat{\alpha}$ parameters


    + Calculate $\mathbf{W}^{*} = [\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}] - \hat{\theta} [\mathbf{x}_{1},\mathbf{x}_{2},\mathbf{z}_{1},\mathbf{z}_{2}]$  
    
    + Calculate $y_{it}^{*} = y_{it} - \hat{\theta} y_{it}$

    + Calculate $\mathbf{V} = [(\mathbf{X}_{1} - \overline{\mathbf{X}}_{1})^{\prime}, (\mathbf{X}_{2} - \overline{\mathbf{X}}_{2})^{\prime}, \mathbf{Z}_1, \overline{\mathbf{X}}_{1} ]$  
    
    + IVR: Regress $y_{it}^{*}$ on $\mathbf{W}^{*}$ with IV set: $\mathbf{V}$ to get $\hat{\beta}$ and $\hat{\alpha}$ parameters.

    + Use robust standard errors. 
    

---- 

## HT Estimation: empirical example

Capturing the real returning of school (on wage) is not an easy task. There are unobserved aspects of ability that are not observed, therefore we would like to run a Random Effects estimator model for Panel Data. Nevertheless, there is a strong correlation between the observed person-speciffic aspects, in this case years of education, and the unobserved ability.


```{r}
colnames(Wages.df)

```


**Variable description**

* exp: years of full-time work experience.

* wks: weeks worked.

* bluecol: blue collar?

* ind: works in a manufacturing industry?

* south: resides in the south?

* smsa: resides in a standard metropolitan statistical area?

* married: married?

* sex: a factor with levels "male" and "female"

* union: individual's wage set by a union contract?

* ed: years of education.

* black: is the individual black?

* lwage: logarithm of wage.


**Regressor classification** 

* $\mathbf{X}_1$: `[wks, south, smsa, married]`  Time-variant, uncorrelated with $\mu$

* $\mathbf{X}_2$: `[exp, exp^2, bluecol, ind, union]`  Time-variant, correlated with $\mu$

* $\mathbf{Z}_1$: `[sex, black]`   Time-invariant, uncorrelated with $\mu$

* $\mathbf{Z}_2$: `[ed]`    Time-invariant, correlated with $\mu$


```{r}
ht <- pht(lwage~wks+south+smsa+married+exp+I(exp^2)
            +bluecol+ind+union+sex+black+ed |
            sex+black+wks+south+smsa+married,
            data=Wages.df, model ="ht", index=c("id","time"))
summary(ht)
```



```{r}
# pht() is deprecated, alternative estimation: 
# .. syntax from plm help: ?plm - does not follow model specification above...
# .. see Baltagi, B.H. (2013) Econometric Analysis of Panel Data, 5th ed., John Wiley and Sons., 
#    chapter 7.4 for detailed discussion.
ht2 <- plm(lwage ~ wks + south + smsa + married + exp + I(exp ^ 2) + 
             bluecol + ind + union + sex + black + ed |
             bluecol + south + smsa + ind + sex + black |
             wks + married + union + exp + I(exp ^ 2),
          data=Wages.df, random.method = "ht", 
          model = "random", inst.method = "baltagi", 
          index=c("id","time"))
summary(ht2)
```
