---
title: "Hierarchical data/models with 3 or more levels"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(knitr)
require(ggplot2)
require(dplyr)
require(nlme)
require(lme4)
require(gridExtra)
require(lattice)
require(RLRsim)
```

## Hierarchical data example

* Level 1) Student
* Level 2) Class  
* Level 3) School
* Level 4) Corporation (public/private entity operating several schools)



## Empirical data

* In this example, we focus on `general reading achievement` of student.


| Variable | Description                                                          |
|----------|----------------------------------------------------------------------|
| school   |  school id (numerical)                                               |
| corp     |  corporation id (numerical)                                          |
| class    |  class id (numerical)                                                |
| geread   |  general reading achievement (score, **higher is better**)           |
| gevocab  |  student's vocabulary test score                                     |
| clenroll |  size of reading class - no. of students                             |
| cenroll  |  school size - no. of students                                       |


```{r}
Achieve <- read.csv("data/Achieve.csv")
Achieve$gender <- as.factor(Achieve$gender)
str(Achieve)
length(unique(Achieve$class)) # note that class numbering is school-specific
length(unique(Achieve$school)) # number of schools
length(unique(Achieve$corp)) # number of corporations
```


--- 

## Simple model, random intercept, 3 levels


* For hierarchical structure, we use `random = ~1|school/class` where `school` is the higher level data group.  


```{r}
M1 <- lme(geread~1, random = ~1|school/class, data = Achieve)
summary(M1)
```

Interpretation closely follows 2-level models:

* Two different sets of random effects:

   + Random effects for `~1|school` model intercept variability across schools  
   + Random effects for `~1|class %in% school` model intercept variability across classrooms within schools 

* At the end of the output, `R` summarizes the sample size for each of the higher level units.

--- 

## Model with added regressors, random intercept, 3 levels

* We add regressors at different levels:

   + `gevocab` is student-specific (Level 1)
   + `clenroll` is class-specific (Level 2)
   + `cenroll` is school-specific (Level 3)
   

```{r}
M2 <- lme(geread~gevocab+clenroll+cenroll, random = ~1|school/class, data = Achieve)
summary(M2)
```

Interpretation: 

* Higher `gevocab` result in higher reading scores  
* The result of `clenroll` is somewhat confusing as larger classrooms would lead to better reading scores.  
* School size is not a statistically significant factor.

----

## Simple model, random intercept, 4 levels


* For hierarchical structure, we use `random = ~1|corp/school/class` where `corp` is the highest level data groupping.

```{r}
M3 <- lme(geread~1, random = ~1|corp/school/class, data = Achieve)
summary(M3)
```


---- 


## Model with added regressors, random intercept, 4 levels

* We add regressors at different levels:

   + `gevocab` is student-specific (Level 1)
   + `clenroll` is class-specific (Level 2)
   + `cenroll` is school-specific (Level 3)
   

```{r}
M4 <- lme(geread~gevocab+clenroll+cenroll, random = ~1|corp/school/class, data = Achieve)
summary(M4)
intervals(M4)
```

---- 


## Model with added regressors, random intercept and "slope", 3 levels

* We add regressors at different levels:

   + `gevocab`  is student-specific (Level 1,`i`)  
   + `clenroll` is class-specific (changes only at Level 2, `j`)  
   + `gender`   is student-specific (Level 1, `i`)  
   + school is at level 3, with `k` index  
   
* Random effect added for intercept and gender (which is binary, so beta coefficient is not a true slope)  

$$
\texttt{geread}_{ijk} = \beta_0 + \beta_1 \texttt{gevocab}_{ijk}
+ \beta_2 \texttt{clenroll}_{jk} + \beta_3 \texttt{gender}_{ijk}
+ u_{0j(k)} + u_{3j(k)} + v_{0k} + v_{3k} + \varepsilon_{ijk}
$$


```{r}
M5 <- lme(geread~gevocab+clenroll+gender, random = ~1+gender|school/class, data = Achieve)
summary(M5)
```


* By comparing random effects for `gender`, we can see that gender-based reading score differences are relatively prominent at the school level (school-by-school differences), while they are relatively negligible at the class level (within a given school).