---
title: "HW6"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(dplyr)
library(spatialreg)
library(mapview)
library(spdep)
```



```{r}
London <- st_read("London.gpkg")
str(London)
```

## Data description

* `"MSOA11CD"          ID`
* `"MSOA11NM"          name`
* `"med_house_price"   median house prices per local spatial unit`
* `"no2"               NO2 air pollution (local)`
* `"POPDEN"            population density (local)`
* `"per_mixed"         percentage of the demographic group on total pop.`
* `"per_asian"         ..`
* `"per_black"         ..`
* `"per_other"         ..`
* `"dist_pubs"         average distance to pubs`
* `"geometry"          geometry of the sf object (polygon)`


--- 

## Location: illustrative example

```{r}
sub.London <- London[c(172, 175, 178, 179, 181, 182), ]
mapview(sub.London)
```

--- 

## Homework 6

Use this specification of the model (expand by spatial dependency)  

**Model specification:**  
`log(med_house_price) ~ log(no2) + log(POPDEN) + `
`   per_mixed + per_asian + per_black + per_other`
`   + log(dist_pubs)`



1. Visualize house prices on a map (infomap)  
2. Use contiguity (common borders) to generate the **C** and **W** matrices  
3. Test house prices for spatial dependency  
4. Estimate spatial lag model (spatial dependency on endogenous variable)  
5. Estimate spatial Durbin model (spatial dependency on endogenous variable and on **X**)  
6. Decide which model fits the data better (LR test)  
7. Calculate marginal effects for the model you choose
8. Proide short interpretation for each of the steps above




```{r}
# Homework 6
```



