---
title: "Semivariogram example - Meuse river sediments"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sp)
library(gstat)
data(meuse)
```

+ This illustration uses the Meuse dataset included in gstat which comprises of four heavy metals measured in the top soil in a flood plain along the river Meuse.  

+ Apparently polluted sediment is carried by the river and mostly deposited close to the river bank.  

+ Heavy metal concentrations are from composite samples of an area of approximately 15 m x 15 m (i.e. each *point* on a map represents a 15m by 15m square).  

```{r}
str(meuse)
```


* *x* is a numeric vector; Easting (m) in Rijksdriehoek (RDH) (Netherlands topographical) map coordinates. EPSG:28991 - Geprojecteerd coördinatensysteem Amersfoort / RD Old in meters, Amersfoort heeft coördinaten (0 m, 0 m).
* *y* is a numeric vector; Northing (m) in RDH coordinates
* *cadmium* topsoil cadmium concentration, mg kg-1 soil ("ppm"); zero cadmium values in the original data set have been shifted to 0.2 (half the lowest non-zero value)  
* *copper*, *lead*, etc. by analogy
* *dist* distance to the Meuse; obtained from the nearest cell in meuse.grid
 

```{r}
# Total area sampled and plotted (in meters)
max(meuse$x) - min(meuse$x)
max(meuse$y) - min(meuse$y)
```

--- 

## Coordinates, `sp` and `sf` formats

* The data frame above is not *spatially aware*.

* Function `coordinates()` promotes the data frame `meuse` into a SpatialPointsDataFrame which knows about its spatial coordinates.

```{r, warning=F}
sp::coordinates(meuse) = ~x+y # ?sp::coordinates
class(meuse)
head(meuse)
proj4string(meuse)
proj4string(meuse) <- CRS("+init=epsg:28992") # ?gstat::variogram
proj4string(meuse)
```

---   

+ Most packages used for spatial statistics and spatial econometrics require **sp** format of spatial data, not the **sf**.  

+ If necessary, the two spatial data formats can be "translated" back and forth:


```{r, warning=F}
# sp --> sf objects
meuse_sf <- sf::st_as_sf(meuse)
class(meuse_sf)
head(meuse_sf)
# sf --> sp objects
meuse <- as(meuse_sf, "Spatial")
class(meuse)
head(meuse)
```

---

#### The remaining part of the example is based on `sp` data.

We can use `sp` plotting functions `spplot()` and `bubble()` as illustrated below (note: the x- and y-axis are the spatial coordinates)


```{r}
spplot(meuse, "zinc",  colorkey = TRUE, main = "zinc concentrations (ppm)")
bubble(meuse, "zinc", col=c("#00ff0088"), main = "zinc concentrations (ppm)")
```


Besides points, we may use "raster" data:

```{r}
data(meuse.grid) # we read-in a grid (map)
coordinates(meuse.grid) = ~x+y
proj4string(meuse.grid) <- CRS("+init=epsg:28992") # ?gstat::variogram
gridded(meuse.grid) = TRUE
class(meuse.grid)
# The above-shown Zinc concentrations seem to follow (inversely) from distances to the river:
#
image(meuse.grid["dist"], main="distance to river (red = 0)", axes=T)
```

---- 

## Semivariogram

Semivariogram is produced in a two-step approach:

1. Calculate the sample variogram. This is done with the `variogram()` function.

2. Fit a model to the sample variogram using `fit.variogram()` function

The `variogram()` function takes two arguments: the first denotes how one or more variables interact spatially, and the second is an SPDF where those variables reside. 

* In our example, we assume that there is a constant mean (no spatial trend) for the variable `log(zinc)`

```{r}
lzn.vgm <- variogram(log(zinc)~1, meuse) # calculates sample variogram values 
lzn.vgm # 15 groups, # obs per group, repr. distance, gamma, isotropy assumed
```


For the `fit.variogram()` function, a sample variogram is the first argument. The second is the model (functional form, parameter settings) to be fit to the sample variogram. 

* For a list of all possible variograms that can be used, call `?gstat::vgm`, and to see graphical properties/characteristics of these models, call  `?gstat::show.vgms`.


```{r}
lzn.fit <- fit.variogram(lzn.vgm, model=vgm(psill=1, model="Sph", range=900, nugget=1)) # fit model
#
# psill = Partial sill = Sill - Nugget (initiall value for fitting psill)
# "Sph" - fit to spherical "model"/curve
# range	- range parameter of the variogram model component
# nugget=1 - basically, adds nugget to model; if missing, nugget is omitted
# 
lzn.fit
plot(lzn.vgm, lzn.fit) # plot the fitted variogram and the observed variogram on the same graph
```

----- 

## Fitted semivariograms depend on assumptions/settings used:

**Gaussian** semivariance wrt distance:

```{r}
lzn.gau <- fit.variogram(lzn.vgm, model=vgm(psill=1, "Gau", range=900, nugget=1)) # fit model
lzn.gau
plot(lzn.vgm, lzn.gau) 
```

**Exponential** semivariance wrt distance:

```{r}
lzn.exp <- fit.variogram(lzn.vgm, model=vgm(psill=1, "Exp", range=900, nugget=1)) # fit model
lzn.exp
plot(lzn.vgm, lzn.exp) 
```


------ 

**Directional (anisotropic)** semivariance wrt distance:

```{r}
vgm3.dir <- variogram(log(zinc)~1, meuse, alpha = c(0, 45, 90, 135))
vgm3.fit.d <- fit.variogram(vgm3.dir, model = vgm(.59, "Sph", 1200, .05, anis = c(45, .4)))
vgm3.fit.d
plot(vgm3.dir, vgm3.fit.d, as.table = T)
```


------

For spatio-temporal semivariogram, see e.g. [tutorial here](http://r-video-tutorial.blogspot.com/2015/08/spatio-temporal-kriging-in-r.html)


