---
title: "Principal Components Regression"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(2)
require(knitr)
require(plotly)
require(ggplot2)
require(pls)
require(ISLR)
require(psych)
```

## Data

```{r}
Hitters <- Hitters
str(Hitters)
sum(is.na(Hitters$Salary))
Hitters <- na.omit(Hitters)
```

## Model/task description

We will use the data included in the Hitters dataset to predict players' **Salary** (our dependent variable) by means of PCR.


## Dataset description


| Variable   | Description                                                                         |   |
|------------|-------------------------------------------------------------------------------------|---|
| AtBat      |  Number of times at bat in 1986                                                     |   |
| Hits       |  Number of hits in 1986                                                             |   |
| HmRun      |  Number of home runs in 1986                                                        |   |
| Runs       |  Number of runs in 1986                                                             |   |
| RBI        |  Number of runs batted in in 1986                                                   |   |
| Walks      |  Number of walks in 1986                                                            |   |
| Years      |  Number of years in the major   leagues                                             |   |
| CAtBat     |  Number of times at bat during his   career                                         |   |
| CHits      |  Number of hits during his career                                                   |   |
| CHmRun     |  Number of home runs during his   career                                            |   |
| CRuns      |  Number of runs during his career                                                   |   |
| CRBI       |  Number of runs batted in during   his career                                       |   |
| Cwalks     |  Number of walks during his career                                                  |   |
| League     |  A factor with levels A and N   indicating player's league at the end of 1986       |   |
| Division   |  A factor with levels E and W   indicating player's division at the end of 1986     |   |
| PutOuts    |  Number of put outs in 1986                                                         |   |
| Assists    |  Number of assists in 1986                                                          |   |
| Errors     |  Number of errors in 1986                                                           |   |
| Salary     |  1987 annual salary on opening day   in thousands of dollars                        |   |
| NewLeague  |  A factor with levels A and N   indicating player's league at the beginning of 1987 |   |


### Encoding of factors into binary variables

```{r}
mat <- Hitters[,-19] # exclude salary
unique(mat$Division) #E W
unique(mat$NewLeague) #A N
unique(mat$League) #A N
# encoding factors 
mat$Division <- as.numeric(mat$Division=="E")
mat$NewLeague <- as.numeric(mat$NewLeague=="A")
mat$League <- as.numeric(mat$League=="A")
mat <- as.matrix(mat)
dim(mat) # Matrix of regressors
```


## Principal Component Analysis



### KMO test

KMO 

* KMO > .9 is marvelous, 
* KMO in the .80s, mertitourious, 
* KMO in the .70s, middling,  
* KMO in the .60s, medicore, 
* KMO in the .50s, miserable, and 
* KMO less than .5, unacceptable... PCA not relevant.

```{r}
psych::KMO(mat)
```


### PCA using the `{stats}` function `prcomp()`.

Note the `center=T` (default) and  `scale = T` arguments 

```{r}
pca <- prcomp(mat, center = T, scale. = T)
summary(pca)
names(pca) # see help for detailed information
```



### Loading vectors phi_1 to phi_4
```{r}
pca$rotation[,1:4]
# Eigenvalues in the first four components are > 1.
pca$sdev
```


---


#### Loading vectors are mutually orthogonal

+ For each of the $z_m$ components, $\sum_{j=1}^p\phi_{jm}^2 = 1$ , and $\phi$ vectors are orthogonal:


(shown for principal components 1 - 10)
```{r}
round(t(pca$rotation[,1:10]) %*% pca$rotation[,1:10], 5)
```

--- 

### PCA plot
```{r}
screeplot(pca, type="lines")
abline(h=1, col="red")
```


### Principal components 1 to 4, individuals 1 to 10 (*Z* vectors)  


```{r}
pca$x[1:10,1:4]
```


### Supporting plot: shows Salary on the plain of the first two components
```{r}
gdta <- data.frame(Salary=Hitters$Salary,PC1=pca$x[,1],PC2=pca$x[,2])
gdta$Name <- row.names(Hitters)
ggplotly(
  ggplot(gdta)+
    geom_point(aes(PC1,PC2,color=Salary))+
    scale_colour_gradient(low="red",high="green")+
    xlab("PC1")+
    ylab("PC2"),
  tooltip = c("Salary")
  )
# 3D plot of the same data
# plot_ly(x=pca$x[,1], y=pca$x[,2], z=Hitters$Salary,type="scatter3d", mode="markers", color=Hitters$Salary)
```




# Principal Component Regression

We may simply do OLS on the PCA-generated principal components:
````{r}
summary(lm(Hitters$Salary ~ pca$x[,1:4]))
```


## However, it is more convenient to use specialized PCR functions: 


e.g. `pcr()` from the `{pls}` package.

```{r}
pcr.fit <- pcr(Salary~., data=Hitters, scale=TRUE,validation="CV")
pcr.fit
summary(pcr.fit)
pcr.cv <- crossval(pcr.fit, segments = 10)
plot(MSEP(pcr.cv), legendpos="topright")
summary(pcr.cv, what = "validation")
```

### We may perform the CV-based selection of components to be used in PCR as:

```{r}
selectNcomp(pcr.fit, method = "onesigma", plot = TRUE)
```


### PCR-estimated model, with `ncomp = 1` 

```{r}
pcr.fit.1 <- pcr(Salary~., data=Hitters, scale=TRUE, ncomp=1)
summary(pcr.fit.1)
loadings(pcr.fit.1)[,1, drop=F] # compare with:  pca$rotation[,1:4] ...above
```


### For quick illustration of features of the PCR-estimated model, we use `ncomp = 4` 


(although CV suggests only one PC for regression)

```{r}
pcr.fit.4 <- pcr(Salary~., data=Hitters, scale=TRUE, ncomp=4)
summary(pcr.fit.4)
loadings(pcr.fit.4)[,1:4] # compare with:  pca$rotation[,1:4] ...above
```


### Fitted values from PCR (regression on PC1 only vs regression PC1 to PC4)

```{r}
pcr.predictions <- predict(pcr.fit, Hitters[1:100,], ncomp=1, type="response")
head(cbind(Hitters$Salary[1:100], pcr.predictions))
predplot(pcr.fit, ncomp = 1, newdata = Hitters[1:100,], asp = 1, line = TRUE)
# You can see that the benefit from adding 3 additional components is ver small:
predplot(pcr.fit, ncomp = 4, newdata = Hitters[1:100,], asp = 1, line = TRUE)
```

---

## Illustration of prediction efficiency of the PCR estimation


To assess PCR vs OLS prediction efficiency, we shall split "Hitters" data.frame  into a train sample (model estimation) and test sample (to calculate and compare Salary predictions) at random, approx. 50/50.


### Split the data to train and test subsamples
```{r}
set.seed(1)
train <- sample(c(TRUE,FALSE), nrow(Hitters),rep=TRUE)
test <- (!train)
```

### Calculate test-sample MSE for OLS
```{r}
ols.fit <- lm(Salary~., data=Hitters, subset=train)
summary(ols.fit)
ols.fitted <- predict(ols.fit, newdata=Hitters[test==T,])
MSE.ols <- mean((ols.fitted - Hitters$Salary[test==T])^2)
```

### Calculate test-sample MSE for PCR (1 and 4 components)
```{r}
set.seed(10)
pcr.fit2 <- pcr(Salary~., data=Hitters, subset=train, scale=TRUE, validation="CV")
pcr.pred2.1.comp <- predict(pcr.fit2, Hitters[test==T, ], ncomp=1, type="response")
MSE.pcr.1.c <- mean((pcr.pred2.1.comp - Hitters$Salary[test==T])^2)
pcr.pred2.4.comp <- predict(pcr.fit2, Hitters[test==T, ], ncomp=4, type="response")
MSE.pcr.4.c <- mean((pcr.pred2.4.comp - Hitters$Salary[test==T])^2)
```

### Compare MSE values
```{r}
MSE.ols # Test MSE for the OLS model
MSE.pcr.1.c # Test MSE for the PCR model, 1 component
MSE.pcr.4.c # Test MSE for the PCR model, 4 components
```



----

**Note:** The train sample / test sample setup as shown here is arbitrary and potentially not-representative. Better evaluation/comparison can be performed through k-Fold Cross Validation.