---
title: "Quantile regression"
subtitle: "Quick repetition and empirical example"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(quantreg)
library(Ecdat)	
BudgetFood <- BudgetFood
BudgetFood$town <- as.factor(BudgetFood$town)
```

--- 


## Quantile regression

+ QR allows for studying the impact of independent variables on different quantiles of dependent variable’s distribution, and thus provides a complete picture of the relationship between $y$ and $x_1, x_2, \dots, x_k$.  

+ Robust to outliers in $y$ observations.

+ Estimation and bootstrap inferences are distribution-free.

---

## Empirical example using `BudgetFood` data

### Data description and visualization

CS data 1980, 23972 households from Spain


| Variable   | Description                                           | 
|------------|-------------------------------------------------------|
| wfood      | percentage of total expenditure which the household has spent on food |
| totexp     | total expenditure of the household |
| age        | age of reference person in the household |
| size       | size of the household |
| town       | size of the town where the household is placed categorised into 5 groups: 1 for small towns, 5 for big ones |
| sex        | sex of reference person (man,woman) |





```{r, warning=F}
str(BudgetFood)
#
# pairs(BudgetFood, col="blue", main="Scatterplots")
#
hist(BudgetFood$wfood, prob=TRUE, col = "blue", border = "black")
lines(density(BudgetFood$wfood))
```

---- 

### Compare OLS and LAD estimation

```{r, warning=F}
# OLS
OLS.est <- lm(wfood ~ log(totexp)+age+size+town+sex, data=BudgetFood)
summary(OLS.est)
# LAD 
LAD.est <- rq(wfood ~ log(totexp)+age+size+town+sex, data=BudgetFood, tau = 0.5)
summary(LAD.est)
```

--- 

### Quantile regression: estimation and plots


```{r, warning=F}
QR.seq <- rq(wfood ~ log(totexp)+age+size+town+sex, data=BudgetFood, 
             tau = 1:9/10)
summary(QR.seq)
# summary plot
plot(summary(QR.seq), ols=T)
```


--- 

### Quantile regression: predicting (conditional) quantiles of the DV


```{r, warning=F}
QR.seq <- rq(wfood ~ log(totexp)+age+size+sex, data=BudgetFood, 
             tau = 1:9/10)
BDGF <- BudgetFood[10,]
fp1 <- quantreg:::predict.rq(QR.seq, newdata=BDGF, stepfun = T)
fp1
plot(fp1~seq(0.1,0.9, by =0.1), main = "Conditional WFOOD Quantile Predictions for household 10", type = "h", lwd = 5, ylab = "wfood", xlab = "tau")
#
BDGF <- BudgetFood[1112,]
fp2 <- quantreg:::predict.rq(QR.seq, newdata=BDGF, stepfun = T)
fp2
plot(fp2~seq(0.1,0.9, by =0.1), main = "Conditional WFOOD Quantile Predictions for household 1112", type = "h", lwd = 5, ylab = "wfood", xlab = "tau" )
```


--- 

### Testing for equal parameters at different quantiles of the dependent variable:


```{r, warning=F}
q30 <- rq(wfood ~ log(totexp)+age+size+sex, data=BudgetFood, tau = 0.3)
q50 <- rq(wfood ~ log(totexp)+age+size+sex, data=BudgetFood, tau = 0.5)
anova(q30, q50)
```

