---
title: "Polynomial and step regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR)
library(ggplot2)
attach(Wage)
Wage <- Wage
agelims <- range(Wage$age)
age.grid <- seq (from=agelims [1], to=agelims [2], length.out = 300)
```

## Polynomial regression

+ Quick recap only -- covered in previous courses  
+ Precursor to splines, smoothers and GAMs (discussed next)  

$$
 y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + \dots  + \beta_d x_i^d + u_i
$$

--- 

## `Wage` data example from `{ISLR}`

We shall focus on a regression model of in the general form:
$$\texttt{Wage}~~\leftarrow~~f(\texttt{Age}).$$

#### Linear regression model / linear function  

```{r}
# Linear model
m1 <- lm(wage~age,data = Wage)
summary(m1)$coefficients
predm1 <- predict(m1, interval = "confidence")
# head(predm1)
plotDF.1 <- cbind(Wage, predm1)
#
# Plot the linear model
#
ggplot(plotDF.1, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="blue", alpha = .35) +
  geom_line( aes(y = fit), colour="black", size = 1) +
  theme_bw()

```


#### Linear regression model / quadratic function in `Age`  

```{r}
# Polynomial regression, d=2
m2 <- lm(wage~age+I(age^2),data = Wage)
summary(m2)$coefficients
# we get the same result using poly() function:
(summary(lm(wage~poly(age,degree = 2, raw = T),data = Wage)))$coefficients # note the raw argument
predm2 <- predict(m2, interval = "confidence")
plotDF.2 <- cbind(Wage, predm2)
#
# Plot the data and fitted values 
#
ggplot(plotDF.2, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="blue", alpha = .35) +
  geom_line( aes(y = fit), colour="black", size = 1) +
  theme_bw()

```


**Quick exercise:**  

+ use the `poly()` function to fit a polynomial regression of order $d=4$.  
+ show coefficients of the fitted model
+ use `ggplot2` functionality to display fitted values and confidence intervals  

```{r}
# Polynomial regression, d=4
# m3 <- lm()


# Plot the data and fitted values 


``` 


--- 

## Step (piecewise-constant) regression 


+ Polynomial regression function imposes a **global** structure on the non-linear function. 

+ Instead, we can use a step function where regressor $x$ is broken into different bins and we fit a different constant for each bin (our regressor is transformed into a ordered categorical variable). 

+ Alternatively, we select adequate (ad-hoc) cutpoints $c_1, c_2, \dots, c_K$ for a continuous regressor $x$ and construct $K+1$ dummy variables:

    + $C_0(X) \quad = \quad 1[X < c_1],$  
    + $C_1(X) \quad = \quad 1[c_1 \leq X < c_2]$  
    + $\cdots$  
    + $C_K(X) \quad \!= \quad 1[c_{K} \leq X]$  
    

```{r}
# step (piecewise constant) regression
Wage$ageCut <- cut(Wage$age,4) # 4 levels, i.e. 3 cut-points, equidistant
#
m4 <- lm(wage~ageCut-1,data = Wage)
summary(m4)$coefficients
predm4 <- predict(m4, interval = "confidence")
plotDF.4 <- cbind(Wage, predm4)
# Plot the data and fitted values 
ggplot(plotDF.4, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr, fill=ageCut), alpha = .35) +
  geom_line( aes(y = fit, color=ageCut), size = 1) +
  theme_bw()
```
 

--- 

## Step (piecewise-linear) regression 

+ We can easily extend the above step regression to piecewise linear models  
+ In our `Wage` example $\texttt{Wage}~~\leftarrow~~f(\texttt{Age})$, let's try piecewise LRM with two cutpoints: $c_1$ and $c_2$:  


$$
y_i = 
    \begin{cases}
        \beta_{01} + \beta_{11} x_i +  u_i ~~~\texttt{if}~~~ x_i <  c_1 ;& \\ ~ & \\
        \beta_{01} + \beta_{11} x_i +  u_i ~~~\texttt{if}~~~ c_1 \leq x_i <  c_2 ;& \\ ~ & \\
        \beta_{03} + \beta_{13} x_i + u_i ~~~\texttt{if}~~~ x_i \geq  c_2 .& 
    \end{cases}
$$




```{r}
# step (piecewise constant) regression
Wage$ageCut3 <- cut(Wage$age,3) # 3 levels, i.e. 2 cut-points, equidistant
X1 <- model.matrix(~Wage$ageCut3-1) # -1 means no "base" level is used
colnames(X1) <- c("ageL1","ageL2","ageL3")
Wage <- cbind(Wage,X1)
#
m5 <- lm(wage~ageCut3+I(age*ageL1)+I(age*ageL2)+I(age*ageL3)-1,data = Wage)
summary(m5)$coefficients # 6 parameters estimated
predm5 <- predict(m5, interval = "confidence")
plotDF.5 <- cbind(Wage, predm5)
# Plot the data and fitted values 
ggplot(plotDF.5, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr, fill=ageCut3), alpha = .35) +
  geom_line( aes(y = fit, color=ageCut3), size = 1) +
  theme_bw()
```


Note how the piecewise regression is non-continuous (zoom in at $c_1$):

```{r,warning=F, echo=FALSE}
# zoom-in using the xlim and ylim arguments
ggplot(plotDF.5, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr, fill=ageCut3), alpha = .35) +
  geom_line( aes(y = fit, color=ageCut3), size = 1) +
  ylim(50,150) +
  xlim(30,50) +
  labs(title = "Zoom-in to previous plot") +
  theme_bw()
```

---

#### Interestingly, as we enforce "continuity" to the piecewise regression line, we actually save degrees of freedom:

$$
y_i = \beta_0 + \beta_1 x_i + \beta_2\, (x_i-c_1)_{+} + \beta_3\, (x_i-c_2)_{+} + u_i  
$$


```{r}
# "Continuous" step (piecewise constant) regression elements:
Wage$AgeBasFun1 <- ifelse(Wage$age > 38.7 , Wage$age-38.7,0) #c_1 = 38.7
Wage$AgeBasFun2 <- ifelse(Wage$age > 59.3 , Wage$age-59.3,0) #c_2 = 59.3
#
m6 <- lm(wage~age+AgeBasFun1+AgeBasFun2, data = Wage)
summary(m6)$coefficients # 4 parameters estimated
predm6 <- predict(m6, interval = "confidence")
plotDF.6 <- cbind(Wage, predm6)
# Plot the data and fitted values 
ggplot(plotDF.6, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="red", alpha = .35) +
  geom_line( aes(y = fit), color="red", size = 1) +
  theme_bw()
```


---



## Step (piecewise-polynomial) regression 


The extention from piecewise linear to piecewise cubic functions is relatively easy:  
(for simplicity, let's use 1 knot $c_1$)


$$
y_i = 
    \begin{cases}
        \beta_{01} + \beta_{11} x_i + \beta_{21} x_i^2 + \beta_{31} x_i^3 + u_i ~~~\texttt{if}~~~ x_i <  c_1 ;& \\ ~ & \\
         \beta_{02} + \beta_{12} x_i + \beta_{22} x_i^2 + \beta_{32} x_i^3 + u_i ~~~\texttt{if}~~~ x_i \geq  c_1 .& 
    \end{cases}
$$


```{r}
# step (piecewise constant) regression
Wage$ageCut2 <- cut(Wage$age,2) # 2 levels, i.e. 1 cut-point
#
#
# Please note we use automatic knot/cut-point selection....
# use ?cut() and "breaks" argument to set knots manually.
#
#
X2 <- model.matrix(~ Wage$ageCut2 - 1) # -1 means no "base" level is used
colnames(X2) <- c("ageLow", "ageHigh")
Wage <- cbind(Wage,X2)
#
#
m7 <-lm(wage~ageCut2+I(age*ageLow)+I((age*ageLow)^2)+I((age*ageLow)^3)
        +I(age*ageHigh)+I((age*ageHigh)^2)+I((age*ageHigh)^3) -1, data = Wage)
summary(m7)$coefficients # 8 parameters
predm7 <- predict(m7, interval = "confidence")
plotDF.7 <- cbind(Wage, predm7)
# Plot the data and fitted values 
ggplot(plotDF.7, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr, fill=ageCut2), alpha = .35) +
  geom_line( aes(y = fit, color=ageCut2), size = 1) +
  theme_bw()
```


#### Again, as we enforce "continuity" to the piecewise cubic regression line, we actually save degrees of freedom:

$$
y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \beta_3 x_i^3 + \beta_4 \, (x_i-c_1)_{+}^3 + u_i  
$$



```{r}
# "Continuous" step (piecewise constant) regression
Wage$AgeBasFunc <- ifelse(Wage$age > 49 , Wage$age-49,0) #c = 49
#
m8 <- lm(wage~poly(age,3,raw = T)+I(AgeBasFunc^3), data = Wage)
summary(m8)$coefficients # 5 parameters estimated
predm8 <- predict(m8, interval = "confidence")
plotDF.8 <- cbind(Wage, predm8)
# Plot the data and fitted values 
ggplot(plotDF.8, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="red", alpha = .35) +
  geom_line( aes(y = fit), color="red", size = 1) +
  theme_bw()
```


--- 


**Quick exercise:**  

(a)  
+ formulate a piecewise-cubic function to fit $\texttt{Wage}~~\leftarrow~~f(\texttt{Age})$.  
+ use 2 knots (cutpoints) ... those are already established $c_1 = 38.7,~c_2 = 59.3$ 
+ start by writing-down the regression equation  
    - use the non-continuous version  
    - enforce continuity (and save d.f. at the same time)  

(b)  
+ Estimate the restricted function (continuity enforced) using `R`  
    + note that basis functions $h_1 = (x_i-c_1)_{+}$ and $h_2 = (x_i-c_2)_{+}$ are already estabished in the `R` code above `[l.191-l.192]`.  
+ use `ggplot2` functionality to display fitted values and confidence intervals  

```{r}
# Polynomial regression, d=4
# m9 <- lm()


# Plot the data and fitted values 

``` 


--- 

Examples partially based on [ISLR textbook and R scripts](http://faculty.marshall.usc.edu/gareth-james/ISL/)
