---
title: "Local regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ISLR)
library(locfit)
```



Local regression at $X = x_0$ 

1. Gather the fraction $s = k/n$ of training points whose $x_i$ \\are closest to $x_0$.  
2. Assign a weight $K_{i0} = K(x_i , x_0)$ to each point in this neighborhood, so that the closest point has the highest weight \\and vice versa. All but the nearest $k$ neighbors get zero weights.  
3. Fit \textit{weighted least squares regression} of $y_i$ on $x_i$ by finding $\beta$ parameters that minimize the following expression:
    $$\min : ~~~ \sum_{i=1}^n K_{i0} (y_i - \beta_0 - \beta_1 x_i)^2 \,.$$  
* Fitted value at $x_0$ is given by $\hat{f}(x_0) = \hat{\beta}_0 + \hat{\beta}_1 x_0\,$.  
    

Local regression: choices to be made

* Weighting function $K$ must be defined (e.g.: $\tfrac{1}{(x_0 - x_i)^2}, x_0 \neq x_i$).  

* Functional form in Step 3 (constant, linear, quadratic, etc.).

* Span $s$ controls flexibility of the fit (can choose using CV).

--- 

In `R`, local regression can be performed using:

+ `loess` (Locally Estimated Scatterplot Smoothing / Local Polynomial Regression Fitting) from `{stats}` package.  
+ using the `{locfit}` package.  

--- 

## `Wage` $\leftarrow$ `Age` example

We use the `locfit()` function from `{locfit}` package to estimate local regression.

+ `nn=0.5` - each neigborhood consists of 50% of nearest observations (w.r.t. total sample)  

+  Kernel weight function defaults to = `kern = "tcub"`, which is the tricube kernel  
    + $K(u)= \frac{70}{81}(1-|u|^3)^3~1(|u|<1)$ where $u=\frac{||x-x_0||}{\text{s.d.}(x)}$  
    + (given diverse scales of measurement, $x$ is standardized)  
    + $1()$ is the Indicator function.  
    
+ `deg=2` ($2^{nd}$ order polynomial) is the default setting.


```{r}
LREG <- locfit(wage ~ lp(age,nn=0.5, deg = 2), Wage)
# summary(LREG)
predLREG <- predict(LREG, newdata = Wage$age, band = "local")
# conf int for plotting
lwr <- predLREG$fit-(2*predLREG$se.fit)
upr <- predLREG$fit+(2*predLREG$se.fit)
# Plot the data and fitted values 
ggplot(Wage, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="red", alpha = .35) +
  geom_line( aes(y = predLREG$fit), color="red", size = 1) +
  theme_bw()
```


---


## Increased LOESS flexibility

+ Can be achieved by making the regression more local.  
+ By setting `nn=0.2`, we only consider 20% of nearest observations   

```{r}

LREG <- locfit(wage ~ lp(age,nn=0.2, deg = 2), Wage)
# summary(LREG)
predLREG <- predict(LREG, newdata = Wage$age, band = "local")
# conf int for plotting
lwr <- predLREG$fit-(2*predLREG$se.fit)
upr <- predLREG$fit+(2*predLREG$se.fit)

# Plot the data and fitted values 
ggplot(Wage, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="green", alpha = .35) +
  geom_line( aes(y = predLREG$fit), color="darkgreen", size = 1) +
  theme_bw()


```


----


## Local regression: slope of the fitted line

* Use `nn=0.2` to ensure flexibility


```{r}
# repeat from previous section
LREG <- locfit(wage ~ lp(age,nn=0.2, deg = 2), Wage)
predLREG <- predict(LREG, newdata = Wage$age, band = "local")
lwr <- predLREG$fit-(2*predLREG$se.fit) # approx. for 95% interval
upr <- predLREG$fit+(2*predLREG$se.fit) # approx. for 95% interval
#
# Produce fitted values at all observed age values
pr.unique <- predict(LREG, newdata = unique(Wage$age))
#
# Produce slope values from the estimation, note the "deriv=1" argument
dLREG <- locfit(wage ~ lp(age,nn=0.2, deg = 2), Wage, deriv=1)
#
# Get slopes at each unique observed age value 
slope <- predict(dLREG, newdata = unique(Wage$age))
years <- unique(Wage$age)
#
#
# Besides fitted values, we can plot how the slope of the fit line changes:
ggplot()+
  geom_line( aes(y = slope, x = years), color="darkgreen", size = 1 )+
  geom_hline(yintercept = c(0), linetype="dashed", color='red', size = 0.7)+
  labs(x = "age", y = "slope of the fitted curve") +
  theme_bw()
#
# Plot "local" slopes at selected ages: 37,52,64
# slope for age = 37
x1 <- years[15]-6
xend1 <- years[15]+6
y1 <- pr.unique[15]-6*slope[15]
yend1 <- pr.unique[15]+6*slope[15]
# slope for age = 64
x2 <- years[51]-6
xend2 <- years[51]+6
y2 <- pr.unique[51]-6*slope[51]
yend2 <- pr.unique[51]+6*slope[51]
#
# slope for age = 52
x3 <- years[10]-6
xend3 <- years[10]+6
y3 <- pr.unique[10]-6*slope[10]
yend3 <- pr.unique[10]+6*slope[10]
#
#
# Note that the lenght of the "slope" is exaggerated for visibility.
#
ggplot(Wage, aes(x = age, y = wage) ) +
  geom_point(col="grey70") +
  geom_ribbon( aes(ymin = lwr, ymax = upr), fill="green", alpha = .35) +
  geom_segment(data = data.frame(x = x1, xend=xend1, y=y1, yend=yend1), 
               aes(x=x, y=y, xend=xend, yend=yend), size = 1, color='red')+
  geom_segment(data = data.frame(x = x2, xend=xend2, y=y2, yend=yend2), 
               aes(x=x, y=y, xend=xend, yend=yend), size = 1, color='red')+
  geom_segment(data = data.frame(x = x3, xend=xend3, y=y3, yend=yend3), 
               aes(x=x, y=y, xend=xend, yend=yend), size = 1, color='red')+
  geom_vline(xintercept=c(37,52,64), linetype="dashed", color='red', size=0.6)+
  geom_line( aes(y = predLREG$fit), color="darkgreen", size = 1) +
  theme_bw()
```




               