---
title: "Generalized linear mixed models (binary response)"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
library(dplyr)
library(nlme)
library(lme4)
library(gridExtra)
library(VGAM)
library(MASS)
library(glmmML)
library(effects)
```
# Medical antenatal care in Bangladesh

Data: Bangladesh Demographic and Health Survey (BDHS)

Our response variable `antemed`  is a binary indicator of whether a woman received antenatal care from a medically-trained provider (a doctor, nurse or midwife) at least once before her most recent live birth. 

The data come from the Bangladesh Demographic and Health Survey (BDHS), a nationally representative cross-sectional survey of women of reproductive age (13-49 years).

The data have a two-level hierarchical structure:  

* 5366 women at level 1,  
* nested within 361 communities at level 2. 

In rural areas a community corresponds to a village, while an urban community is a neighbourhood based on census definitions.


| Variable | Description                                                                                                                  |
|----------|------------------------------------------------------------------------------------------------------------------------------|
| comm     |  Community identifier                                                                                                        |
| wonid    |  Woman identifier                                                                                                            |
| antemed  |  Received antenatal care at least   once from a medically-trained provider, e.g. doctor, nurse or midwife  (1 = yes, 0 = no) |
| bord     |  Birth order of child (ranges from   1 to 13)                                                                                |
| mage     |  Mother's age at the child's birth   (in years)                                                                              |
| urban    |  Type of region of residence at   survey (1=urban, 0=rural)                                                                  |
| meduc    | Mother's level of education at survey (1=none, 2=primary, 3=secondary or   higher)                                           |
| islam     |  Mother's religion (1=Islam,   0=other)                                                                                      |
| wealth   | Household wealth index in quintiles (1=poorest to 5=richest)                                                                 |


 
 

```{r}
BDHS <- read.table("_dataW08/BDHS.txt", header = TRUE)
BDHS$comm <- factor(BDHS$comm)
str(BDHS)
mean(BDHS$antemed) # Overall proportion of antenatal care indicator
``` 
 
 

* At **level 1 (individual level)**, we consider variables such as a woman's education and age at the time of birth of the child. 
* **Level 2 (regional level)** variables include an indicator of whether the region of residence is classified as urban or rural. 


### First, we fit a model using only intercept and comunity effect: 

For gaussian response variable, we might use linear mixed effects (LME) model such as:

$E(y_{ij}|u_i) = \beta_0 + u_{0i} + \varepsilon_{ij}, ~~~ u_{0i} \sim N(0, \sigma^2_u) ~~~ \varepsilon_{ij} \sim N(0, \sigma^2)$


Given our binary response variable `antemed`, logistic regression model is used:


$\textit{logit}(p_{ij}) = \log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg) = \beta_0 + u_{0i} + \varepsilon_{ij}$

* $\beta_0$ is the overall intercept for all comunities,
* $u_{0i}$ is the random effect for the $i$-th community. 



```{r}
fit <- glmer(antemed ~ (1 | comm), family = binomial("logit"), data = BDHS)
```

```{r}
summary(fit)
```


$\hat{\beta}_0 = 0.148$. Intercept for $i$-th community is: $0.148 + \hat{u}_{0i}$, variance of $u_{0i}$ is 1.464

The likelihood ratio statistic for testing the null hypothesis that variance of $u_{0i}$ is 0 can be done as two model comparision: by comparing the two-level model, with the corresponding single-level
model (without the level 2 random effect).

```{r}
fit0 <- glm(antemed ~ 1, family=binomial("logit"),data=BDHS)
test <- -2*logLik(fit0) + 2*logLik(fit)
pchisq(test,df=1,lower.tail=FALSE)
```

### Random effects can be extracted and plotted:

```{r}
library(lattice)
qqmath(ranef(fit, condVar=TRUE),strip=FALSE)$comm
```

### Now we include maternal age (`mage`) as an explanatory variable in the model.

First we center `mage`, then estimate the model using the centered value `magec`  

* This approach simplifies interpretation of $\beta_0$ and $u_0$ elements. 

```{r}
mean(BDHS$mage)
BDHS$magec <- BDHS$mage - mean(BDHS$mage)
fit2 <- glmer(antemed~magec+(1|comm),family=binomial("logit"),data=BDHS)
summary(fit2)
```


Note that there is little change in the estimate of the between-community variance `1.462` (as compared to the previous model: `1.464`), suggesting that the distribution of maternal age is similar across communities.

--- 


```{r}
# Plot of predicted probabilities of success (given "magec")
predprob <- predict(fit2, type="response")
predlogit <- logit(predprob)
datapred=unique(data.frame(cbind(predprob = predprob, comm = BDHS$comm,magec = BDHS$magec)))
xyplot(predprob ~ magec, data = datapred, groups = comm, type = "l", col = "darkgreen", xlim = c(-10, 30), ylim = c(0, 1))
```

---

On the scale of the linear predictor, the estimated model can be described as follows: 

$\widehat{\textit{logit}(p_{ij})} = \widehat{\log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg)} = 0.145 -0.032 \textit{magec}  + \hat{u}_{0i}$

```{r}
summary(ranef(fit2)$comm$`(Intercept)`)
```


For a woman of average age (`23.6`), the log-odds of receiving antenatal care range from about -2.1 to 2.4 depending on which community she lives in. This translates to a range in probabilities of $\frac{\exp^{(-2.1)}}{1+\exp^{(-2.1)}}=0.11~$ to $~\frac{\exp^{(2.4)}}{1+\exp^{(2.4)}}=0.92$, so there are strong community effects.

---

### Now, we include two additional variables,`wealth` and `meduc`.

We assume fixed effects in the new variables (random effect only on the intercept)

```{r}
BDHS$meduc <- factor(BDHS$meduc)
BDHS$wealthc <- BDHS$wealth-mean(BDHS$wealth) # we center wealth
fit3 <- glmer(antemed~magec+wealthc+meduc+(1|comm), family=binomial("logit"), data=BDHS)
summary(fit3)
```



**Variance Partition Coefficient**

The variance partition coefficient (VPC) is calculated as $\frac{\sigma^2_u}{\sigma^2_u + \sigma^2}$. 
From theory, the standard logistic distribution has variance $\sigma^2 = \frac{\pi^2}{3} = 3.29$. Then:

```{r}
as.numeric(summary(fit3)$varcor$comm)
```

and:

$$  \frac{\sigma^2_u}{\sigma^2_u + \sigma^2} = \frac{0.867}{0.867+3.29}=0.208$$

Thus 21% of the residual variation in the propensity to use antenatal care services is attributable to unobserved community characteristics.





**Predict probabilities from a Multilevel model **

$$p_{ij} = \frac{e^{z_{ij}}}{1+e^{z_{ij}} }$$
where 

$$ z_{ij} = -0.490 - 0.005 \textit{magec}_{ij} + 0.403 \textit{welthc}_{ij} + 0.545 \textit{meduc2}_{ij}+  1.305 \textit{meduc3}_{ij}$$

For example, given model `fit3` and community $i$, for a woman of mean age `(magec=0)`, and no education `(meduc2=meduc3=0)`, the forecast is:

$$ \widehat{log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg)} = (-0.490 + \hat{u}_i) + 0.403\, \textit{wealthc}_{ij}$$

Also, fitted values $\hat{p}_{ij}$ (`type="response"`) can be produced as


```{r}
head(predict(fit3, type="response")) # fitted "success" probabilities.
```




## Mixed effects (two-level) model with random slope

In the previous model, maternal age, education and household wealth are assumed to be the same in each community. We will now relax this assumption and fit a random slope (coefficient) for wealth to allow its effect to vary across communities.


$$ log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg) = \beta_0 + \beta_1 \textit{magec}_{ij} + \beta_3 \textit{meduc}_{ij} + \beta_4 \textit{welthc}_{ij} + \beta_5 \textit{wealthc}_{ij} + u_{0i} + u_{4i} \textit{wealthc}_{ij} + \varepsilon_{ij} $$
Note that the new term $u_{4i}$ has been added to the model, so that the coefficient
of `wealthc` has become $\beta_{4i} = \beta_4 + u_{4i}$

```{r}
fit4 <- glmer(antemed~magec+meduc+wealthc + (1|comm) + (0+wealthc|comm), data = BDHS, family=binomial("logit"))
summary(fit4)
```

Note that we specified the random term `(0+wealthc|comm)` which implies that the random intercepts and slopes are not correlated.

Now, we can test whether the generalization from model `fit3` is valid (i.e. adding random slope improves model properties):

```{r}
anova(fit3,fit4) # do not reject H0
```

Hence, we disregard mode `fit4` and use `fit3` results.




## We can include a Level 2 explanatory variable `urban`

( `urban` varies among communities, not across individuals)

```{r}
BDHS$urban <- factor(BDHS$urban)
levels(BDHS$urban) <- c("rural","urban")
fit5 <- glmer(antemed~magec+meduc+wealthc+urban+(1|comm), 
              data = BDHS, family=binomial("logit"))
summary(fit5)
```

There exists significant differences between rural and urban regions of residence.
We find that urban women are more likely than rural women to use antenatal care.



***



# Deer Data Assigment 

### Data from Zurr - Mixed Effects Models and Extensions in Ecology with R

Relationship between the presence and absence of larvae (Elaphostrongylus cervi) in deer across Spain and the explanatory variables `Length` the deer and the `Farm`.


| Variable | Description                                                                                                                  |
|----------|------------------------------------------------------------------------------------------------------------------------------|
| infect    |  Binary dependent variable: presence (1) and absence (0) of larvae                                                 |
| Length    |  size of the deer                                                                                                           |
| Farm       |  Farm ID                                                                                                           |

```{r}
deer <- read.table('_dataW08/deer.txt',header=TRUE)
deer$Farm <- factor(deer$Farm)
str(deer)
ggplot(deer) + geom_boxplot(aes(as.factor(infect), Length))
ggplot(deer) + geom_density(aes(Length, group=as.factor(infect), fill= as.factor(infect)),alpha=1/5)+
   theme_bw()
```

## Assignment

Estimate the following models, print & interpret the summary.


**1) Estimate model with fixed effects**

* Produce a centered value of Lenght: `Lenghtc`
* Estimate model:  infect ~ Fixed (1+Lenghtc+Farm)

```{r}

```


**2) Estiamte model with random intercept**

* infect ~ Fixed (1+Lenghtc) + Random (intercept, changing along Farm)

```{r}

```


**3) Estiamte a model with random slope and intercept**

* infect ~ Fixed (1+Lenghtc) + Random (intercept+Lenghtc, changing along Farm)

```{r}

```



### Plot of predicted probabilities of parasitic infection. 

The thick line in the middle represents the predicted values for the ‘population of farms’
(approx. validity only)

```{r, warning=FALSE}
deer.glmm <- glmer(infect~Length+(1|Farm),data=deer,family=binomial)
deer.glm <- glm(infect~Length,data=deer,family=binomial)
ggplot(deer) + 
  geom_smooth(aes(Length,fitted(deer.glmm), group=Farm, color=Farm),
              method="loess")+
  geom_smooth(aes(Length,fitted(deer.glm)),
              size=2, 
              color="black")+
  theme_bw()
```





Based on lecture by [Dae-Jin Lee](http://idaejin.github.io/bcam-courses/neiker-2016/material/mixed-models/) and slightly modified. [cc](https://creativecommons.org/licenses/by-sa/3.0/)