---
title: "Manipulating and analyzing data with dplyr"
output: html_document
---


------------

> ### Learning Objectives
>
> * Select certain columns in a data frame with the **`dplyr`** function `select`.
> * Filter certain rows in a data frame according to filtering conditions with the **`dplyr`** function `filter` .
> * Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`.  
> * Add new columns to a data frame that are functions of existing columns with `mutate`.  
> * Use `summarize` and `group_by` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results.

----

# Data Manipulation using **`dplyr`**

* Bracket subsetting `[,]` (with logical operators) is handy, but it can be cumbersome and difficult to read, especially for complicated operations. 

* **`dplyr`** is a package for making tabular data manipulation easier. 

* It pairs nicely with **`tidyr`** package and functionality.

* To learn more about **`dplyr`** and **`tidyr`** , you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf) and this [one about **`tidyr`**](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf).


```{r, echo=F, warning=F, message = F}
## load dplyr
library("dplyr")
```

## Sample data


```{r, message = FALSE,}
## load sample data
NUTS2.DF <- read.csv("datasets/NUTS2data.csv")
# summary(NUTS2.DF)
str(NUTS2.DF)
```

| Variable    | Description                                                     |
|-------------|-----------------------------------------------------------------|
| Year        |  time identification of the observation 2010 - 2016             |
| NUTS2       |  NUTS2 geographic identification of the observation             |
| NUTS0       |  State-level identification (AT BE CZ DE DK HU LU NL PL SI SK)  |
| GDP_MIO_EUR |  GDP in Mio EUR per NUTS2 per Year                              |
| TotPopNr    |  Number of inhabitants                                          |
| Area        |  geographic area in km sq.                                      |


--- 

# Basic **`dplyr`** functionality

## `select()`  - subset columns

* The first argument to this function is the data frame: `NUTS2.DF`,
* Subsequent arguments are the columns to keep.

```{r}
DF2 <- select(NUTS2.DF, Year, NUTS2, TotPopNr)
head(DF2,10)

```

* To select all columns *except* certain ones, put a "-" in front of the variable to exclude it.

```{r}
DF3 <- select(NUTS2.DF, -TotPopNr, -Area)
head(DF3,10)

```


--- 

## `filter()`  - subset rows on conditions

* To choose rows based on a specific criteria, use `filter()`:



```{r, purl = FALSE}
# Choose all records for Slovenia
filter(NUTS2.DF, NUTS0 == "SI")
```



```{r, purl = FALSE}
# Choose all records for both Slovenia AND year 2011
filter(NUTS2.DF, Year == 2011, NUTS0 == "SI")
# Same year (2011), retrieve data for Slovenia and Slovakia
filter(NUTS2.DF, Year == 2011, NUTS0 %in% c("SI","SK"))
```

---


## Pipes `%>%`

What if you want to select and filter at the same time? There are three
ways to do this: 

+ use intermediate steps, 
+ nested functions, 
+ pipes.

--- 

#### Sample exercise: 

* Get `TotPopNr` data (*plus id info*, i.e. columns `Year` and `NUTS2`) for Slovenia, year 2011 and older: 

With **intermediate steps**, you create a temporary data frame and use
that as input to the next function, like this:

```{r}
DF4 <- filter(NUTS2.DF, Year <= 2011, NUTS0 == "SI")
DF5 <- select(DF4, Year, NUTS2, TotPopNr)
DF5
```

* This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

--- 

You can also **nest functions** (i.e. one function inside of another), like this:

```{r}
select(filter(NUTS2.DF, Year <= 2011, NUTS0 == "SI"), Year, NUTS2, TotPopNr)
```

* This is handy, but can be difficult to read if too many functions are nested, as `R` evaluates the expression from the inside out (in this case, filtering is done first, then selecting).

--- 

The last option, **pipes**, are a recent addition to `R`. Pipes let you take
the output of one function and send it directly to the next, which is useful
when you need to do many things to the same dataset.  

Pipes in R look like `%>%` and are available within `{dplyr}` (as well as other packages). 

* If you use RStudio, you can type the pipe with <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC 
* or <kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r}
NUTS2.DF %>%
  filter(Year <= 2011, NUTS0 == "SI") %>%
  select(Year, NUTS2, TotPopNr)
```

* The pipe operator `%>%` takes the object on its left and passes it as the **first argument to the function on its right**,  
* we don't need to explicitly include the data frame as an argument to the `filter()` and `select()` functions any more.

* In the above code, we use the pipe to send the `NUTS2.DF` dataset first through `filter()` to keep rows where `Year` is $\leq$ 2011 AND `NUTS0 == "SI"`, then through `select()` to keep only the `Year`, `NUTS2` and `TotPopNr` columns. 
    * In our example, you can reverse the steps and still get the same results. 
    * However, this is not a general case & the **ordering of steps matters!**

* Some may find it helpful to read the pipe like the word **then**. 

* The **`dplyr`** functions by themselves can be somewhat simple, but by combining them into linear workflows with the pipe, we can accomplish complex manipulations of data frames.

* Many types of functions (from different packages) can be used in pipes.  

* If we want to create a new object using the pipe, we do it the usual way: we use `<-` operator and assign a name to the new object:

```{r}
NUTS.SI <- NUTS2.DF %>%
  filter(Year <= 2011, NUTS0 == "SI") %>%
  select(Year, NUTS2, TotPopNr)
NUTS.SI # display the new objec (df) 
```

* Note that the final data frame is the leftmost part of the piping expression.

--- 

**Quick exercise 1:** 

Complete `R` script below as follows, based on the `NUTS2.DF` data frame: 

* Retrieve GDP data for Austria: filter for Austria (`"AT"`), Czechia (`"CZ"`) AND Year 2015,
* Select columns: NUTS2, GDP_MIO_EUR,
* Use the pipe syntax

```{r}
# Uncomment and complete the task
#NUTS2.DF %>%

```

---

## `mutate()`  - create new columns using information in other columns

```{r}
# For Slovenia, calculate GDP per capita (in EUR)
NUTS2.DF %>%
  filter(NUTS0 == "SI") %>%
  mutate(GDPpc = (GDP_MIO_EUR/TotPopNr)*1000000) %>% # conversion to EUR from mioEUR
  head(10) # pipes work with non-dplyr commands as well (if dplyr is loaded)
```

--- 

**Quick exercise 2:** 

Complete `R` script below as follows: 

Show population dentisty by NUTS2 regions  

* Use  year 2016 only (filter),
* Calculate population density `PopDens` = (TotPopNr / Area) ,
* Show columns: NUTS2, `PopDens`,
* Use the pipe syntax
* Show the first 15 rows in your Rmd output,

```{r}
# Uncomment and complete the task
#NUTS2.DF %>%

```

--- 

## `group_by()` and `summarize()`  - summary on grouped data

* `group_by()` takes an existing tbl and converts it into a grouped tbl where operations are performed "by group". `ungroup()` removes grouping.  

* For example, we can use GDP per capita at the NUTS2 level and calculate state-wise (NUTS0) averages.

```{r}
# Calculate average value of GDP per capita at the State level, year 2016
# .. serves for illustration only - NUTS2 to NUTS0 averages are not weighted by population
NUTS2.DF %>%
  filter(Year == 2016) %>%
  mutate(GDPpc = (GDP_MIO_EUR/TotPopNr)*1000000) %>%
  group_by(NUTS0) %>% 
  summarize(mean_GDPpc = mean(GDPpc, na.rm = TRUE)) 
```

* Note the output is not a `data.frame` table, but a `tibble` - `{dplyr}` / `{tidyverse}` specific format.

* Many data analysis tasks can be approached using the *split-apply-combine* paradigm: split the data into groups, apply some analysis to each group, and then combine the results. 

* `group_by()` is often used together with `summarize()`, which collapses each group into a single-row summary of that group.  `group_by()` takes as arguments the column names that contain the **categorical** variables for which you want to calculate the summary statistics.

--- 

## `group_by()` and `mutate()`  

* May be used for calculations on grouped data, 
* Easy to calculate lags and individual means for panel data

```{r}
# For Slovenia and Slovakia, calculate first lag of GDP and individual means (over time) for TotPopNr
SISK <- NUTS2.DF %>%
  select(-Area) %>% 
  filter(NUTS0 %in% c("SI","SK")) %>%
  group_by(NUTS2) %>% 
  mutate(GDP_lag1 = lag(GDP_MIO_EUR, k = 1), PopAvg = mean(TotPopNr)) %>% 
  ungroup() %>% 
  arrange(NUTS2,Year) %>% 
  select(Year,NUTS2,GDP_MIO_EUR,GDP_lag1,PopAvg)
head(as.data.frame(SISK),20)
```


--- 

## `arrange()`  - sort results

```{r}
# For Slovenia, calculate first lag of GDP and sort: fist by region, then by time
NUTS2.DF %>%
  select(-Area, -TotPopNr) %>% 
  filter(NUTS0 == "SI") %>%
  group_by(NUTS2) %>% 
  mutate(GDP_lag1 = lag(GDP_MIO_EUR, k = 1)) %>% 
  ungroup() %>% 
  arrange(NUTS2,Year) # sorts by NUTS2, then by Year - both ascending
```

* To sort in descending order, use `desc()`.
* e.g. `arrange(desc(NUTS2),Year)`
* You can use `ungroup()` in the pipe for removing the grouping (e.g. for subsequent analysis).


---- 

# Joining data from multiple datasets

* Here, we start by reading-in some additional dataset 

| Variable    | Description                                                     |
|-------------|-----------------------------------------------------------------|
| Year        |  time identification of the observation 2011 - 2016. Compared to `NUTS2.DF`, there is  **no 2010**   daata  |
| NUTS2       |  NUTS2 id   **(same 113 regions) **                                 |
| Unem        |  Unemployment rate in %                                         |

```{r, message = FALSE,}
## load sample data
Unem <- read.csv("datasets/NUTS2data2.csv")
str(Unem)
```

--- 

## `left_join()`  - joins two datasets

##### Start with `NUTS2.DF` and *append* `Unem` dataset

```{r, message = FALSE,}
# Note the missing 2010 Unem values
NewDF <- left_join(NUTS2.DF, Unem, by = c("Year", "NUTS2"))
str(NewDF)
# Show output - head of the table only
NewDF %>% 
 arrange(NUTS2,Year) %>% 
 head(12)
```

* All observations in the `left` dataset (`NUTS2.DF`) are preserved.

* `NA` generated if `Unem` observation for a given `"Year", "NUTS2"` combination is not available

--- 

Now, we join the two dataframes again - using `left_join()`, but using **alternative ordering od data.frames to join:**  

* Start with `Unem` (on the left) and *append* `NUTS2.DF` dataset.  
* The `left_join()` command works easily within the pipe operator sequence.  

```{r, message = FALSE,}
# Show output - head of the table only
Unem %>% # in the pipe, Unem becomes the first argument to left_join()
  left_join(NUTS2.DF, by = c("Year", "NUTS2")) %>% 
  arrange(NUTS2,Year) %>% 
  head(12)
```

* Note the changed ordering of columns.

* All observations in the `left` dataset (`Unem`) are preserved.

* Observations for `2010` in `NUTS2.DF` are NOT *imported*, as there is no such combination of `"Year", "NUTS2"` in the `Unem` dataset (i.e. on the left).

* `inner_join()` , `right_join()` are available in `dplyr` package.

* Joining by columns of different names:  
    + Say, `Year` variable is in both DFs, but `NUTS2` region-IDs of the DF1 are in a corresponding `geo` column of the second DF:  
    + `left_join(DF1,DF2, by = c("Year", "NUTS2"="geo"))`  
    + With pipes, `DF1` is implicit, comes from the previous "line" of the pipe
    

* Sometimes, `merge()` from the `base` package may be a reasonable alternative to `join` commands from `dplyr`.  

---

**Quick exercise:**

* Start with the `NUTS2.DF` dataset
* Filter for year 2015 & NUTS2 regions in Germany (`"DE"`)
* For the NUTS2 regions selected (filtered), append all corresponding data from the `Unem` dataset
* Create new dataframe: save ID column(s), Unemployment and GDP data into a new DF
* Arrange output by NUTS2 ID
* Use pipe operator syntax, show the output 


```{r}

```

--- 

This worksheet draws from 

* [Manipulating, analyzing and exporting data with tidyverse](https://github.com/datacarpentry/R-ecology-lesson/blob/master/03-dplyr.Rmd) 