---
title: "Eurostat database and package"
output: html_document
---

### `{eurostat}` 

`{eurostat}` R package provides tools to access open data from [Eurostat](http://ec.europa.eu/eurostat): 

* data search, 
* download, 
* manipulation, 
* visualization. 


---  

## Using the package

* [Cheat sheet: eurostat R package](https://raw.githubusercontent.com/rstudio/cheatsheets/main/eurostat.pdf)
* [Tutorial (vignette) for the eurostat R package](https://ropengov.github.io/eurostat/articles/eurostat_tutorial.html)
* [Detailed documentation for eurostat functions](http://ropengov.github.io/eurostat/index.html)

--- 

| Command              | Description                                          |  
|----------------------|------------------------------------------------------|  
| `get_eurostat_toc()` | Download table of contents of Eurostat datasets      |  
| `search_eurostat()`  | Retrieve datasets titles from Eurostat (regex syntax)     |  
| `get_eurostat()`     | Read Eurostat data                                   |  
| `label_eurostat()`   | Get Eurostat code descriptions                       |  
| `get_eurostat_geospatial()`   | Download geospatial data from GISGO         |  


--- 

## Search Eurostat for data

```{r, echo=T, message=F, warning=F}
options(readr.default_locale=readr::locale(tz="Europe/Berlin"))
library(eurostat)
library(sf)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(zoo)
```


```{r, eval=FALSE}
# To actually run this {r} chunk, change the eval argument
toc <- get_eurostat_toc() # Downloads Table of Contents of Eurostat Data Sets
class(toc)
dim(toc)
str(toc,list.len = 10) # only few items listed
```

With `search_eurostat()`, you can search the table of contents for particular text (text patterns). 

* *regex*: R regular expression syntax is used: see `?regex` for details.

* `.*` is particularly useful basic "tool" in text pattern search: 

    + The period `.` matches any single character.  
    + `*`: The preceding item (`.`) will be matched zero or more times.  

* **regex is case sensitive** -- see next example, where we search Eurostat for unemployment data:


```{r}
# kable() generates tabular (formatted) output in Rmd files
# .. fixed=F allows for complex/flexible matching, vs the default "as is" setting
kable(search_eurostat(".*unemployment.*rates.*NUTS", fixed=F))
kable(search_eurostat(".*Unemployment.*rates.*NUTS", fixed=F))
```

```{r, eval=FALSE, echo=T}
# Alternatively, you can use grep() to search a downloaded TOC
# .. this way, you can deal with the case-sensitivity "issue"
toc <- get_eurostat_toc() # Downloads Table of Contents of Eurostat Data Sets
toc[grep(".*unemployment.*rates.*NUTS", toc$title, ignore.case = T),]$title
# ... this R code is not executed, provided for your information only
# ... you can switch the `eval` argument to produce the output table
```


--- 


## Download data

As an example, let's choose the **Unemployment rates by sex, age and NUTS 2 regions (%)** dataset	`lfst_r_lfu3rt`

* All datasets are available through a web browser
* https://ec.europa.eu/eurostat/databrowser/view/lfst_r_lfu3rt/default/table?lang=en

Download the data:

```{r, message=F}
# https://ec.europa.eu/eurostat/databrowser/view/lfst_r_lfu3rt/default/table?lang=en
Un.DF <- get_eurostat("lfst_r_lfu3rt", time_format = "num") # note the simplified time format
dim(Un.DF)
str(Un.DF)
UnemploymentDF <- Un.DF %>% 
  mutate(unit = as.factor(unit),
         age = as.factor(age),
         sex = as.factor(sex),
         geo = as.factor(geo),
         isced11 = as.factor(isced11))
summary(UnemploymentDF) # note the "values" variable - it contains Unemployment rates for a given "row"
```


By default, variable identification is provided through Eurostat **codes**.

* To get human-readable labels instead, we can use `label_eurostat()` function
* Labels are good for orientation in the dataset, NOT for wider/longer data handling and other data manipulations.


```{r, echo=T, message=F}
Un.DF.l <- label_eurostat(Un.DF, fix_duplicated = T)
```

```{r}
head(Un.DF.l,6)
```

Also, codes and their "descriptions" can be shown side-by-side:

```{r, warning=F}
kable(cbind(as.character(unique(Un.DF$isced11)),as.character(unique(Un.DF.l$isced11))))
kable(cbind(as.character(unique(Un.DF$age)),as.character(unique(Un.DF.l$age))))
kable(cbind(as.character(unique(Un.DF$sex)),as.character(unique(Un.DF.l$sex))))
kable(cbind(as.character(unique(Un.DF$geo)),as.character(unique(Un.DF.l$geo)))[1:17,])
# note the NUTS-code format:
# NUTS0 (states) have 2-digit IDs ... "AT"
# NUTS1 regions have 3-digit IDs
# NUTS2 regions have 4-digit IDS 
#
```

---

## Data handling

We can simply save the data for subsequent use:

 * To perform the operation, disable the `eval=FALSE` argument.

```{r, eval=FALSE}
write.csv(Un.DF, "datasets/Unemployment.csv", row.names = F)
```

----

We can use `{tidyverse}` and `{ggplot2}` functions to filter and plot data. 

### **Example 1: Unemployment plot for selected countries (long-format data)**  

Using the `Un.DF`, perform the following data handling (mostly filtering):  

* `Y15-74` i.e. age group *from 15 to 74 years*  
* Years 2010 and newer  
* Total unemployment only (no M/F/T) structure  
* All education levels combined
* Select only: Austria, Czech Republic, Germany, Hungary, Poland, Slovakia  
* Show NUTS0 leve only (State-level)  



```{r}
Un.DF %>% 
  filter(age == "Y15-74", 
         TIME_PERIOD >= 2010, 
         # sex == "T", 
         isced11 == "TOTAL",
         geo %in% c("AT","CZ","DE","HU","PL","SK")) %>% 
  ggplot(aes(x = TIME_PERIOD, y = values, colour = geo))+ # plot filtered data
    geom_line()+ # choose plot type
    ggtitle("Unemployment rates")+ # Define main title
    ylab("Unemployment (%)")+  # define label on the y-axis
    theme_minimal()  # choose plot "design"
```

--- 

### Quick exercise: 

**Total wages/compensations per region by NACE r2 activities**

* Use pipe operator for steps 4 & 5.  
* (you may find it easier to perform this exercise in a separate R-script, load `eurostat` `ggplot2` and `dplyr` first)  

1. Download "nama_10r_2coe" from Eurostat (use simplified time format).  
2. Check the structure of your dataset. 
3. Find out the meaning of `currency` and `nace_r2` codes.  
4. Retrieve data for the sector "Financial and insurance activities" for CZ, SK, SI, AT (NUTS0), years 2007-2017, in euro.
5. Plot the data using `ggplot` (follow formatting in Example 1 or choose your own style)

* This is just a simple exercise focused on downloading and filtering data from Eurostat... data among states are not directly comparable (we would need a per-employee standardization).

```{r}
#
#
```

---

### **Example 2: Choropleth (infomap) of unemployment**

We continue with the previously used `Un.DF` Unemployment dataset as in **Example 1** (i.e. not with data from Quick exercise).  

The following tasks (filtering) will be performed:  

* `Y15-74` i.e. age group *from 15 to 74 years*  
* Use data for 2017  
* Total unemployment only (no M/F/T)  
* All education levels
* Use NUTS2-level data
* Use only the following countries: Austria, Czech Republic, Germany, Hungary, Poland, Slovakia  
* Draw choropleth (cartogram, infomap)

```{r, echo=T, message=F}
# Download geospatial data from GISCO
geodata <- get_eurostat_geospatial(nuts_level = "2")
# Filter Unemployment dataset for the year 2017
Un.2017 <- Un.DF %>% 
  filter(age == "Y15-74", 
         TIME_PERIOD == 2017, 
         sex == "T", 
         isced11 == "TOTAL") %>% 
  filter(nchar(as.character(geo)) == 4) %>% # NUTS2 regions have a 4-digit id
  mutate(NUTS0 = substr(as.character(geo), start=1, stop=2)) %>% # retrieve NUTS0 id from NUTS2
  filter(NUTS0 %in% c("AT","CZ","DE","HU","PL","SK")) 
# Join Unemployment data with "map data"
map_data <- inner_join(geodata, Un.2017)
# plot the data
ggplot()+ 
  geom_sf(data = map_data, aes(fill = values))+ 
  # note that "values" is name of column that stores unemployment data...
  scale_fill_gradientn('Unem. \n(%) ', colours=brewer.pal(8, "Reds"))+
  ggtitle("Unemployment rates")
```

---

**Quick assignment:** Add Netherlands (`NL`) to the infomap.

--- 


### **Example 3: Save Unemployment time series data into a `zoo` object **


#### Long format (not a TS object)


```{r}
Un.DF %>% 
  filter(age == "Y15-74", 
         TIME_PERIOD >= 2010, 
         sex == "T", 
         isced11 == "TOTAL",
         geo %in% c("AT","CZ","DE","HU","PL","SK")) %>% 
  head(10)
```


#### TS format (`zoo`-type TS object)


```{r}
# Step 1 - wider format
Unempl <- Un.DF %>% 
  filter(age == "Y15-74", 
         TIME_PERIOD >= 2010, 
         sex == "T", 
         isced11 == "TOTAL",
         geo %in% c("AT","CZ","DE","HU","PL","SK")) %>% 
  pivot_wider(names_from = "geo", values_from = "values") # convert to wider format
head(Unempl) # wider format, not a TS/zoo object
#
# Step 2:  convert to zoo format
# Here, I choose to drop columns that have zero variability 
# (given the filtering that was performed in previous steps)
# plus the time column (as information is contained in the zoo object)
Unempl.zoo <- zoo(Unempl[,-c(1:5)], order.by = Unempl$TIME_PERIOD) 
class(Unempl.zoo)
head(Unempl.zoo)
```

---

#### Quarterly data - {eurostat} to `zoo` - example (TS)

* Second example on transforming Eurostat data into TS using '{zoo}'  
* This time, we use quarterly data (for monthly data, steps are analogous)  


First, we download a new dataset that contains quarterly data: `lfsq_urgan`

[Unemployment rates by sex, age and citizenship (%)](https://ec.europa.eu/eurostat/databrowser/view/lfsq_urgan/default/table?lang=en)

```{r}
Un.Q <- get_eurostat("lfsq_urgan", time_format = "num") # note the simplified time format
str(Un.Q)
Un.Q %>% 
  mutate(unit = as.factor(unit),
         age = as.factor(age),
         sex = as.factor(sex),
         citizen = as.factor(citizen)) %>% 
  summary()
# Plot unemployment of foreigners (%)
Un.Q <- Un.Q %>% 
  filter(age == "Y15-74", 
         TIME_PERIOD >= 2010, 
         sex == "T", 
         citizen == "FOR",
         geo %in% c("AT","CZ","DE")) %>%
  pivot_wider(names_from = "geo", values_from = "values") # convert to wide format
head(as.data.frame(Un.Q)) # wide format, not a TS object
# Note the "order.by = as.yearqtr" syntax
# .. I drop redundant columns
Un.Q.zoo <- zoo(Un.Q[,-c(1:5)], order.by = as.yearqtr(Un.Q$TIME_PERIOD)) 
Un.Q.zoo$TIME_PERIOD <- NULL
head(Un.Q.zoo)
# Simple TS plotting from the {ggplot2} package
autoplot(Un.Q.zoo) 
```
--- 


