---
title: "TS imputation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(imputeTS)
```

## Univariate Time Series Missing Value Imputation

---

### This example is based on the `{imputeTS}` package.

**`tsAirgap` dataset** has 144 rows and the incomplete version includes 13 NA values. The series represents
monthly totals of international airline passengers from 1949 to 1960 (in thousands).

#### Visualize the missing values in this time series

```{r}
class(tsAirgap)
summary(tsAirgap)
head(tsAirgap,24)
imputeTS::ggplot_na_distribution(tsAirgap) # {imputeTS} function, based on ggplot2
```


#### The next plot is especially useful if the time series would otherwise be too large to be plotted.  

+ Multiple observations for time intervals are grouped together and represented as bars.   
+ For these intervals, information about the amount of missing values are shown  

```{r}
ggplot_na_distribution2(tsAirgap)
```

#### Print stats about the missing data

```{r}
statsNA(tsAirgap)
```

--- 

### Different types of imputation performed using the `{imputeTS}` package.

#### **Missing value imputation by mean value** 

```{r}
impMean.Airgap <- na_mean(tsAirgap)
ggplot_na_imputations(tsAirgap, impMean.Airgap) 
#
#  In this illustrative example, we have the original (full) TS
#  and we can use it to evaluate different imputation methods....
#
ggplot_na_imputations(tsAirgap, impMean.Airgap, tsAirgapComplete) #  tsAirgapComplete is the complete dataset
```

---

#### **Imputation by last observation carried forward (LOCF)**

```{r}
impLOCF.Airgap <- na_locf(tsAirgap)
ggplot_na_imputations(tsAirgap, impLOCF.Airgap, tsAirgapComplete)
```


---

#### **Imputation by interpolation: linear & spline**

#### Linear interpolation

```{r}
impLinear.Airgap <- na_interpolation(tsAirgap) # Linear interpolation
ggplot_na_imputations(tsAirgap, impLinear.Airgap, tsAirgapComplete)
```

#### Spline interpolation

```{r}
impSpline.Airgap <- na_interpolation(tsAirgap, option = "spline") # Spline interpolation
ggplot_na_imputations(tsAirgap, impSpline.Airgap, tsAirgapComplete)
```

--- 

#### **Kalman filter -- for TS with trend & seasonality**

#### Kalman filter

+ The Kalman filter is an efficient recursive filter estimating the internal-state of a linear dynamic system from a series of noisy measurements. 

```{r}
impKalman.Airgap <- na_kalman(tsAirgap, model = "StructTS", smooth = TRUE)
ggplot_na_imputations(tsAirgap, impKalman.Airgap, tsAirgapComplete)
```

--- 

## Multivariate Time Series Missing Value Imputation

Package `{Amelia}` supports time series MI:


 + IP (Imputation & Posterior distribution-based method), EM (Expectation maximisation algorithms)  
    + `Amelia` combines **PMM** techniques and **MVN** distribution modelling  
 + We may include time polynomials up to third order as additional variables in the covariance matrix.  
 + Another option is using time lagged variables (leads, lags).  
 + Bootstrap approach is used to account for the variance of parameters in a function used to produce imputations.  
 + For details, see e.g. https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2996611
 + or https://gking.harvard.edu/amelia
 
---

#### **Amelia -- TS example**

Based on the Freetrade dataset: *Economic and political data on nine developing countries in Asia from 1980 to 1999. This dataset includes 9 variables*:  

* year,  
* country,  
* average tariff rates,  
* polity IV score,  
* total population,  
* gross domestic product per capita,  
* gross international reserves,  
* dummy variable for if the country had signed an IMF agreement in that year,  
* a measure of financial openness,  
* a measure of US hegemony.  

 
```{r, message=F}
library(Amelia)
summary(freetrade) 
missmap(freetrade) # description of missing data in "freetrade" data
#
# Say, we are interested in estimating this type of model:
# .. complete cases estimation used
summary(lm(tariff ~ polity + pop + gdp.pc + year + country,
          data = freetrade))
```


+ When performing multiple imputation, the first step is to identify the variables to include in the imputation model.

    * `a.out <- amelia(freetrade, m = 5, ts = "year", cs = "country")`

+ Many variables that are recorded over time within a cross-sectional unit are observed to vary smoothly over time. In such cases, knowing the observed values of observations close in time to any missing value may enormously aid the imputation of that value. However, the exact pattern may vary over time within any cross-section. 

+ Amelia can build a general model of patterns within variables across time by creating a sequence of polynomials of the time index.  

+ If, for example, tariffs vary smoothly over time, then we make the modeling assumption that there exists some polynomial that describes the economy in cross-sectional unit $i$ at time $t$ as ($3^{rd}$ order polynomial is a maximum order that can be used in Amelia):  
$$ y_{it} = \beta_0 + \beta_1 t + \beta_2 t^2 + \beta_3 t^3 + u_{it}$$

---  

As we combine MI and *third-order polynomial time trend*, we may do MI as follows:  

* All available series are used for MI (ID variables - TS and CS - have to be provided)  
* `polytime = 2` is used by default - so this argument has to be set to 3 for a 3rd order polynomial.  
* The estimated time-dynamics coefficients apply to all CS units (i.e. are NOT CS-specific -- this will be relaxed next).   
* `p2s` controls calculation details displayed  

Note:
* MI is applied to all series with missing data (Multiple imputation involves imputing m values for each missing cell, observations remain unchanged),  
* In this example, we mostly focus on the  `tariff` variable


```{r}
set.seed(100)
a.out <- amelia(freetrade, m=5, ts = "year", cs = "country", polytime = 3, p2s = 2) # EM: expectation maximization (ML-base method)
summary(a.out)
tscsPlot(a.out, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out, var = "tariff") #not specific to Malaysia
plot(a.out) # overall basic Imputation information plot

# compare original dataset vs 1st imputation (of the total 5 imputations)
head(freetrade,20) # original data
head(a.out$imputations$imp1,20) # impuation No. 1
```


---  

#### We can impute TS with trends that are specific to each cross-sectional unit:

* `intercs` is a logical argument indicating if the time effects of *polytime* should vary across the cross-section,  
* we use 2nd order polynomials here.  


```{r}
set.seed(120)
a.out.time <- amelia(freetrade, m=5, ts = "year", cs = "country", polytime = 2,
                     intercs = TRUE, p2s = 0) # "p2s = 0" turns off text output
tscsPlot(a.out.time, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out.time, var = "tariff") #not specific to Malaysia
plot(a.out.time) # overall basic Imputation information plot
```
 
 + Note how the imputed values fit original observations better.
 
 
---  

#### We can add leads and lags to calculate the imputation (along with trends specific to each cross-sectional unit):

* leads and lags are only used for the `tariff` series (user-defined setting is applied, can be amended)

* `lags` - a vector of numbers or names indicating columns in the data that should have their lags included in the imputation model. (leads by analogy)

```{r}
set.seed(140)
a.out2 <- amelia(freetrade, ts = "year", cs = "country", lags = "tariff",
                 leads = "tariff", intercs = TRUE, polytime = 2, p2s = 0)
tscsPlot(a.out2, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out2, var = "tariff") #not specific to Malaysia
plot(a.out2) # overall basic Imputation information plot
```
 
 + Calculation time for the last imputation task increases considerably.


#### Estimation can be repeated with the imputed data:

```{r}
b.out <- NULL
se.out <- NULL
for(i in seq_len(a.out2$m)) {
  ols.out <- lm(tariff ~ polity + pop + gdp.pc + year + country, data = a.out2$imputations[[i]])
  b.out <- rbind(b.out, ols.out$coef)
  se.out <- rbind(se.out, coef(summary(ols.out))[, 2])
}
# Not as nice as the {mice} package functionality, yet sufficient:
combined.results <- mi.meld(q = b.out, se = se.out)
combined.results
round(cbind(t(combined.results$q.mi),t(combined.results$se.mi)),5)
```


---


Additional MI approaches (using prior information, etc.) and evaluation procedures can be observed from:

[Enders, C.K., Applied Missing Data Analysis (2022)](https://www.appliedmissingdata.com/)

https://www.jstatsoft.org/article/view/v045i07  

https://cran.r-project.org/web/packages/Amelia/vignettes/amelia.pdf  

https://cran.r-project.org/web/packages/Amelia/vignettes/using-amelia.html  

https://www.ece.iastate.edu/~namrata/EE527_Spring08/Dempster77.pdf  (EM algortihm)
