---
title: "TS imputation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(imputeTS)
```

## Univariate Time Series Missing Value Imputation

---

### This example is based on the `{imputeTS}` package.

**`tsAirgap` dataset** has 144 rows and the incomplete version includes 13 NA values. It represents
the monthly totals of international airline passengers from 1949 to 1960. The

#### Visualize the missing values in this time series

```{r}
class(tsAirgap)
summary(tsAirgap)
head(tsAirgap,24)
plotNA.distribution(tsAirgap)
```


#### The next plot is especially useful if the time series would otherwise be too large to be plotted.  

+ Multiple observations for time intervals are grouped together and represented as bars.   
+ For these intervals, information about the amount of missing values are shown  

```{r}
plotNA.distributionBar(tsAirgap, breaks = 20)
```

#### Print stats about the missing data

```{r}
statsNA(tsAirgap)
```

--- 

### Different types of imputation performed using the `{imputeTS}` package.

#### **Missing value imputation by mean value** 

```{r}
impMean.Airgap <- na_mean(tsAirgap)
plotNA.imputations(tsAirgap, impMean.Airgap) #  tsAirgapComplete is the complete dataset
plotNA.imputations(tsAirgap, impMean.Airgap, tsAirgapComplete) #  tsAirgapComplete is the complete dataset
```

---

#### **Imputation by last observation carried forward (LOCF)**

```{r}
impLOCF.Airgap <- na_locf(tsAirgap)
plotNA.imputations(tsAirgap, impLOCF.Airgap, tsAirgapComplete)
```


---

#### **Imputation by interpolation: linear & spline**

#### Linear interpolation

```{r}
impLinear.Airgap <- na_interpolation(tsAirgap) # Linear interpolation
plotNA.imputations(tsAirgap, impLinear.Airgap, tsAirgapComplete)
```

#### Spline interpolation

```{r}
impSpline.Airgap <- na_interpolation(tsAirgap, option = "spline") # Spline interpolation
plotNA.imputations(tsAirgap, impSpline.Airgap, tsAirgapComplete)
```

--- 

#### **Kalman filter -- for TS with trend & seasonality**

#### Kalman filter

```{r}
impKalman.Airgap <- na_kalman(tsAirgap, model = "StructTS", smooth = TRUE) # Spline interpolation
plotNA.imputations(tsAirgap, impKalman.Airgap, tsAirgapComplete)
```

--- 

## Multivariate Time Series Missing Value Imputation

Package `{Amelia}` supports time series MI:

 + We may include time polynomials up to third order as additional variables in the covariance matrix.  
 + Another option is using time lagged variables (leads, lags).  
 + Bootstrap approach is used to account for the variance of paramters in a function used to produce imputations.  
 + For details, see e.g. https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2996611
 
---

#### **Amelia -- TS example**

Based on the Freetrade dataset: *Economic and political data on nine developing countries in Asia from 1980 to 1999. This dataset includes 9 variables*:  
+ year,  
+ country,  
+ average tariff rates,  
+ polity IV score,  
+ total population,  
+ gross domestic product per capita,  
+ gross international reserves,  
+ dummy variable for if the country had signed an IMF agreement in that year,  
+ a measure of financial openness,  
+ a measure of US hegemony.  

 
```{r}
library(Amelia)
summary(freetrade) 
missmap(freetrade) # description of missing data in "freetrade" data
```


+ Amelia can build a general model of patterns within variables across
time by creating a sequence of polynomials of the time index.  

+ If, for example, tariffs vary smoothly over time, then we make the modeling assumption that there exists some polynomial that describes the economy in cross-sectional unit $i$ at time $t$ as ($3^{rd} order polynomial is a maximum):  
$$ y_{it} = \beta_0 + \beta_1 t + \beta_2 t^2 + \beta_3 t^3 + u_{it}$$

---  

For a second-order polynomial and the `tariff` series, we may do MI as follows:

```{r}
set.seed(100)
a.out <- amelia(freetrade, ts = "year", cs = "country", polytime = 2)
tscsPlot(a.out, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out, var = "tariff") #not specific to Malaysia
```


---  

We can also impute with trends specific to each cross-sectional unit:

```{r}
set.seed(120)
a.out.time <- amelia(freetrade, ts = "year", cs = "country", polytime = 2,
                     intercs = TRUE, p2s = 0) # "p2s = 0" turns off text output
tscsPlot(a.out.time, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out.time, var = "tariff") #not specific to Malaysia
```
 
 + Note how the imputed values fit original observations better.
 
 
---  

We can add leads and lags to calculate the imputation (along with trends specific to each cross-sectional unit):

```{r}
set.seed(140)
a.out2 <- amelia(freetrade, ts = "year", cs = "country", lags = "tariff",
                 leads = "tariff", intercs = TRUE, polytime = 2, p2s = 0)
tscsPlot(a.out2, cs = "Malaysia", main = "MI with Amelia",
         var = "tariff", ylim = c(-10, 60))
compare.density(a.out2, var = "tariff") #not specific to Malaysia
```
 
 + Calculation time for the last imputation task increases considerably.
 


---


Additional MI approaches (using prior information, etc.) and evaluation procedures can be observed from `Amelia` vignette: 

https://cran.r-project.org/web/packages/Amelia/vignettes/amelia.pdf 


